# Intergration Pipeline within seurat to integrate data from both batched of GP33+ scRNA/TCR-seq at various timepoints after infection

# load common libraries and data
```{r}
suppressPackageStartupMessages({
  library(tidyverse) 
  library(Seurat)
  library(RColorBrewer)
  library(viridis)
  library(cowplot)
  library(ggridges)
  library(data.table)
  library(ggpubr)
  library(ComplexHeatmap)
})

path <- "~/ChronicMemoryGithub_Submission/1_scRNATCRseqAtlas"
outs <- paste0(path, "/01_FilterMultiseqIntegration/outs")

so.list.cr119 <- read_rds(paste0(path, "/data/1A_DemuxedSeuratObjectList_batch2.rds"))
so.list.cr082 <- read_rds(paste0(path, "/data/1B_DemuxedSeuratObjectList_batch1.rds"))
```

# remove genes function
```{r}
#remove TCR genes from VF lists
remove_genes <- function(features, species, tcr=TRUE, ig=TRUE, cell_cycle=TRUE, mito=TRUE, histone=TRUE, ribosome=TRUE) {
    print(paste("Before filtering:", length(features), "features"))

    if (species == "human") {
        if (tcr) {
            features <- features[!grepl("^TR[ABGD][VDJC]", features)]
        }
        if (ig) {
            features <- features[!grepl("^IG[HLK][VDJC]", features)]
        }
        if (cell_cycle) {
            features <- features[!(features %in% unlist(cc.genes))]
        }
        if (mito) {
            features <- features[!grepl("^MT-", features)]
        }
    }
    if (species == "mouse") {
        if (tcr) {
            features <- features[!grepl("^Tr[abgd][vdjc]", features)]
        }
        if (tcr) {
            features <- features[!grepl("^Tcr[abgd][vdjc]", features)]
        }
        if (ig) {
            features <- features[!grepl("^Ig[hlk][vdjc]", features)]
        }
        if (cell_cycle) {
            features <- features[!(features %in% stringr::str_to_title(unlist(cc.genes)))]
        }
        if (mito) {
            features <- features[!grepl("^Mt-", features)]
        }
      
       if (histone) {
            features <- features[!grepl("^Hist", features)]
       }
      
       if (ribosome) {
            features <- features[!grepl("^Rps", features)]
       }
      
      if (ribosome) {
            features <- features[!grepl("^Rpl", features)]
        }
      
    }

    print(paste("After filtering:", length(features), "features"))
    
    return(features)
}
```

# rename metadata
```{r}
# select only GP33+ cells from batch 1 (CR82) 
so.list.cr082.gp33 <-  list(so.list.cr082[[1]], so.list.cr082[[2]], so.list.cr082[[3]])

so.list.cr082.gp33 <- so.list.cr082.gp33 %>% lapply(function(so){ 
 so@meta.data <- so@meta.data %>% mutate(group = paste(orig.ident, multiseq_class, sep = "_")) %>%
    separate(group, into = c("Tet", "LaneID", "Mouse")) %>%
    mutate(Tet = substr(Tet, 0,4)) %>%
    mutate(Sort = "Total") %>%
    mutate(LCMV = ifelse(grepl( "Chron", Mouse), "Cl13", NA)) %>% mutate(LCMV = ifelse(grepl("Naive", Mouse), "Naive", LCMV)) %>%
        mutate(LCMV = ifelse(grepl("Acu", Mouse), "Arm", LCMV)) %>%
    mutate(DPI = ifelse(grepl("Chron", Mouse) & !grepl("ChronMem", Mouse), "d26", NA)) %>% mutate(DPI = ifelse(grepl("ChronMem", Mouse), "d100", DPI)) %>%
      mutate(DPI = ifelse(grepl("Acu", Mouse), "d69", DPI)) %>%
    mutate(REP = ifelse(Mouse != "Naive" & Mouse != "Doublet", substr(Mouse, nchar(Mouse), nchar(Mouse)), NA)) %>%
    mutate(Group = ifelse(LCMV == "Naive", "Naive", paste(LCMV, DPI, Tet, sep = "_"))) %>%
    mutate(Group = ifelse(LCMV == "Naive", "Naive", paste(LCMV, DPI, Tet, Sort, sep = "_"))) %>% select(-Mouse) %>% select(-LaneID) %>% 
    mutate(orig.ident = "CR82")
  return(so)
  })

# select only gp33+ cells from batch 2

so.list.cr119.gp33 <- list(so.list.cr119[[1]], so.list.cr119[[3]], so.list.cr119[[4]], so.list.cr119[[5]])

so.list.cr119.gp33 <- so.list.cr119.gp33 %>% lapply(function(so){ 
  so@meta.data <- so@meta.data %>% mutate(group = paste(orig.ident, multiseq_class, sep = "_")) %>% 
    separate(group, into = c("LaneID", "Tet", "Sort", "LCMV", "DPI", "REP")) %>% 
    select(-LaneID) %>% 
    mutate(Group = ifelse(LCMV == "Naive", "Naive", paste(LCMV, DPI, Tet, sep = "_"))) %>%
    mutate(Group = ifelse(LCMV == "Naive", "Naive", paste(LCMV, DPI, Tet, Sort, sep = "_"))) %>% 
    mutate(orig.ident = "CR119")
  return(so)
  })

# merge everything together into a new list
so.list <- c(so.list.cr119.gp33[[1]], so.list.cr119.gp33[[2]], so.list.cr119.gp33[[3]], so.list.cr119.gp33[[4]],
            so.list.cr082.gp33[[1]], so.list.cr082.gp33[[2]], so.list.cr082.gp33[[3]]) 


```

# pre-process seurat object
```{r}
so.list <- lapply(so.list, function(x) {
  # remove doublets and unidentified cells from individual so w/in so.list
  x <- subset(x, !is.na(multiseq_class) & multiseq_class != "Doublet")
  x <- NormalizeData(x)
  x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
  return(x)
})
```

# integration workflow: reciprocal PCA
```{r}

## find overlapping features
### remove TCR genes etc
features <- SelectIntegrationFeatures(object.list = so.list) %>% 
  remove_genes(species = "mouse" , tcr=T, ig=F, cell_cycle=F, mito=F, histone=F, ribosome=F)

## PCA run on integration features, not all features
so.list <- lapply(so.list, FUN = function(x) {
    x <- ScaleData(x, features = features, verbose = F)
    x <- RunPCA(x, features = features, verbose = F)
})

## ID anchors
anchors <- FindIntegrationAnchors(object.list = so.list, anchor.features = features, reduction = "rpca")#, k.anchor = 5)

## created integrated dataset
so.rpca <- IntegrateData(anchorset = anchors)

## specify that we will perform downstream analysis on the corrected data note that the
### original unmodified data still resides in the 'RNA' assay
DefaultAssay(so.rpca) <- "integrated"

```

# scale data and add cell cycles
```{r}
so.rpca <- ScaleData(so.rpca)
so.rpca <- RunPCA(so.rpca, npcs = 30, features = VariableFeatures(so.rpca)) # variable features were set to the cleaned DEG

# add cell cycle
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes
so.rpca <- CellCycleScoring(so.rpca, s.features = s.genes, g2m.features = g2m.genes)
```

# first pass of clustering
```{r}
so.clusters.1 <- RunUMAP(so.rpca, dims = 1:12)

so.clusters.1@meta.data <- so.clusters.1@meta.data %>% 
  mutate(Group = ifelse(LCMV == "Arm", "Arm", NA)) %>% 
  mutate(Group = ifelse(LCMV %in% c("Cl13", "Chron") & DPI == "d100", "Cl13 d100+", Group)) %>% 
  mutate(Group = ifelse(LCMV %in% c("Cl13", "Chron") & DPI != "d100", "Cl13 d21-26", Group)) %>% 
  mutate(Group = ifelse(LCMV == "Naive", "Naive", Group))


DefaultAssay(so.clusters.1) <- "integrated"
so.clusters.1 <- FindNeighbors(so.clusters.1, dims = 1:12)
so.clusters.1 <- FindClusters(so.clusters.1,resolution = 0.5)

DimPlot(so.clusters.1, label = T)
DimPlot(so.clusters.1, split.by = "Group", ncol = 2) & NoAxes()

DefaultAssay(so.clusters.1) <- "RNA"
FeaturePlot(so.clusters.1, features = c("Klrg1", "Cx3cr1", "Pdcd1", "Tox", "Cd101" , "Tcf7", "Sell", "Cxcr3"), ncol = 4) & NoLegend() & NoAxes()
DimPlot(so.clusters.1, split.by = "Group", ncol = 2, group.by = "orig.ident", shuffle = T) & NoAxes()


FeaturePlot(so.clusters.1, features = c("percent.mt", "nCount_RNA", "nFeature_RNA")) & NoAxes()
VlnPlot(so.clusters.1, features = c("percent.mt", "nCount_RNA", "nFeature_RNA"))
DimPlot(so.clusters.1, group.by = "Phase") & NoAxes()

# calculate DEG
cluster.markers <- FindAllMarkers(so.clusters.1, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25, assay = "RNA", features = features)

# get just the top 5 featurs per cluster
cluster.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> topgenes

hm.df <- DotPlot(so.clusters.1, features=unique(topgenes$gene))$data[, c("features.plot", "id", "pct.exp", "avg.exp.scaled")]
## mutate the data bit
hm.df <- hm.df[,c(1,2,4)] %>% pivot_wider(names_from = id, values_from = avg.exp.scaled)%>% as.data.frame
rownames(hm.df) <- hm.df$features.plot

## make heatmap
hm <- hm.df[,-1] %>% 
  Heatmap(name = "Z-score", # name over the legend
          col = rev(brewer.pal(9, "RdBu")), # values for the cell values
          height = unit(16, "in"), width = unit(3, "in"), 
          show_column_dend = F, show_row_names = T, column_dend_height = unit(0.15, "in"), 
          cluster_rows = T, cluster_columns = T # do you want the rows and columns to cluster
        )

# pdf(file = paste0(outs, "/1C_Clusters_1.pdf"), height = unit(18, "in"), width = unit(6, "in"))
hm
# dev.off()

DimPlot(so.clusters.1, label = T)
# ggsave("1C_UMAP_1.pdf", path = outs)

```


# second pass of clustering
## remove junk cells and naive cells
```{r}
# set bad clusters
### c6 = naive
### c13 = histones
### c14 = likely doublets -  express KLRG1 and Sell
### c15 = high mito - high nfeature

so.clusters.1$keep <- "yes"
so.clusters.1$keep[so.clusters.1$seurat_clusters %in% c("6", "13", "14", "15")] <- "no"
so.clusters.1$keep[so.clusters.1$multiseq_class == "Naive"] <- "no"
DimPlot(so.clusters.1, group.by = "keep")

# susbet seurat obj 
so.clusters.2 <- subset(so.clusters.1, keep == "yes")

# seurat processing
DefaultAssay(so.clusters.2) <- "integrated"
so.clusters.2 <- FindVariableFeatures(so.clusters.2)

features.2 <- remove_genes(VariableFeatures(so.clusters.2), 
                         species = "mouse" , tcr=T, ig=F, cell_cycle=T, mito=T, histone=T, ribosome=T)

so.clusters.2 <- ScaleData(so.clusters.2, verbose = T)
so.clusters.2 <- RunPCA(so.clusters.2, npcs = 30, verbose = T, features = features.2)

# clustering & UMAP
so.clusters.2 <- FindNeighbors(so.clusters.2, dims = 1:12)
so.clusters.2 <- FindClusters(so.clusters.2,resolution = 0.5)
so.clusters.2 <- RunUMAP(so.clusters.2, dims = 1:12, return.model = T)

# find DEG
cluster.markers <- FindAllMarkers(so.clusters.2, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25, assay = "RNA", features = features)

cluster.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> topgenes
hm.df <- DotPlot(so.clusters.2, features=unique(topgenes$gene))$data[, c("features.plot", "id", "pct.exp", "avg.exp.scaled")]
hm.df <- hm.df[,c(1,2,4)] %>% pivot_wider(names_from = id, values_from = avg.exp.scaled)%>% as.data.frame
rownames(hm.df) <- hm.df$features.plot

hm <- hm.df[,-1] %>% 
  Heatmap(name = "Z-score", # name over the legend
          col = rev(brewer.pal(9, "RdBu")), # values for the cell values
          height = unit(16, "in"), width = unit(3, "in"), 
          show_column_dend = F, show_row_names = T, column_dend_height = unit(0.15, "in"), 
          cluster_rows = T, cluster_columns = T # do you want the rows and columns to cluster
        )

# pdf(file = paste0(outs, "/1C_Clusters_2.pdf"), height = unit(18, "in"), width = unit(6, "in"))
hm
# dev.off()

DimPlot(so.clusters.2, label = T)
# ggsave("1C_UMAP_2.pdf", path = outs)

DimPlot(so.clusters.2, split.by = "Group", ncol = 3) & NoAxes()

DefaultAssay(so.clusters.2) <- "RNA"
FeaturePlot(so.clusters.2, features = c("Klrg1", "Cx3cr1", "Pdcd1", "Tox", "Cd101" , "Tcf7", "Sell", "Cxcr3"), ncol = 4) & NoLegend() & NoAxes()
VlnPlot(so.clusters.2, features = c("percent.mt", "nCount_RNA", "nFeature_RNA"), pt.size = 0) 

```

# third pass of dataset -  remove junk cells 
```{r}
# set bad clusters
## c11 = high MHC-2
## c10 = still look naive

so.clusters.2$keep <- "yes"
so.clusters.2$keep[so.clusters.2$seurat_clusters %in% c("11", "10")] <- "no"
so.clusters.2$keep[so.clusters.2$multiseq_class == "Naive"] <- "no"
DimPlot(so.clusters.2, group.by = "keep")

# susbet seurat obj 
so.clusters.3 <- subset(so.clusters.2, keep == "yes")

# seurat processing
DefaultAssay(so.clusters.3) <- "integrated"
so.clusters.3 <- FindVariableFeatures(so.clusters.3)

features.3 <- remove_genes(VariableFeatures(so.clusters.3), 
                         species = "mouse" , tcr=T, ig=F, cell_cycle=T, mito=T, histone=T, ribosome=T)

so.clusters.3 <- ScaleData(so.clusters.3, verbose = T)
so.clusters.3 <- RunPCA(so.clusters.3, npcs = 30, verbose = T, features = features.3)


```

# cluster cells with 12 dims
```{r}
DefaultAssay(so.clusters.3) <- "integrated"

# clustering & UMAP
so.clusters.3 <- FindNeighbors(so.clusters.3, dims = 1:12)
so.clusters.3 <- FindClusters(so.clusters.3,resolution = 0.4)
so.clusters.3 <- RunUMAP(so.clusters.3, dims = 1:12, return.model = T)

# find DEG
cluster.markers <- FindAllMarkers(so.clusters.3, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25, assay = "RNA", features = features.3)

cluster.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> topgenes
hm.df <- DotPlot(so.clusters.3, features=unique(topgenes$gene))$data[, c("features.plot", "id", "pct.exp", "avg.exp.scaled")]
hm.df <- hm.df[,c(1,2,4)] %>% pivot_wider(names_from = id, values_from = avg.exp.scaled)%>% as.data.frame
rownames(hm.df) <- hm.df$features.plot

hm <- hm.df[,-1] %>% 
  Heatmap(name = "Z-score", # name over the legend
          col = rev(brewer.pal(9, "RdBu")), # values for the cell values
          height = unit(16, "in"), width = unit(3, "in"), 
          show_column_dend = F, show_row_names = T, column_dend_height = unit(0.15, "in"), 
          cluster_rows = T, cluster_columns = T # do you want the rows and columns to cluster
        )

# pdf(file = paste0(outs, "/1C_Clusters_3.pdf"), height = unit(18, "in"), width = unit(6, "in"))
hm
# dev.off()

DimPlot(so.clusters.3, label = T)
# ggsave("1C_UMAP_3.pdf", path = outs)

DimPlot(so.clusters.3, split.by = "Group", ncol = 3) & NoAxes()
DimPlot(so.clusters.3, split.by = "Group", group.by = "orig.ident",ncol = 3, shuffle = T) & NoAxes()

DefaultAssay(so.clusters.3) <- "RNA"
FeaturePlot(so.clusters.3, features = c("Klrg1", "Cx3cr1", "Pdcd1", "Tox", "Cd101" , "Tcf7", "Sell", "Cxcr3"), ncol = 4) & NoLegend() & NoAxes()
VlnPlot(so.clusters.3, features = c("percent.mt", "nCount_RNA", "nFeature_RNA"), pt.size = 0) 
```

# save RDS
```{r}
saveRDS(so.clusters.3, file = paste0(path, "/99_seurat_objects/1C_IntegratedSeuratObject.rds"))
```

