# Assess DEGs between clusters and look at cluster distribtuion per group

## load common libraries + load SO + set paths
```{r}
suppressPackageStartupMessages({
  library(tidyverse) 
  library(Seurat)
  library(RColorBrewer)
  library(viridis)
  library(cowplot)
  library(ggridges)
  library(ggrepel)
  library(data.table)
  library(ComplexHeatmap)
  library(ggpubr)
  library(ComplexHeatmap)
  library(circlize)
  library(writexl)
})

path <- "~/ChronicMemoryGithub_Submission/1_scRNATCRseqAtlas"
so <- readRDS(paste0(path, "/data/1C_IntegratedSeuratObject.rds"))
outs <- paste0(path,"/output")
```

## set aesthetics for plots
```{r}
# palette
clusterpal <- c(
  "Tcm" = "#65c7ca",
  "Tem_1" = "#f8a523",
  "Tem_2" = "#f6d13a",
  "Tex-Term" = "#b31e0e",
  "Tex-Int" = "#de311e",
  "Tex-KLR" = "#E27E73",
  "Tex-ISG" = "#a0a1d1",
  "Tex-Prog" = "#a94d9a",
  "Prolif" = "#e6af7c"
  )

# plot themes
plot.theme.umap <- theme_cowplot() + theme(plot.title = element_text(hjust = 0.5), 
        #panel.border = element_rect(colour = "black", fill=NA, linewidth = 0.5, color = "grey80"),
        axis.text = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5, face = "plain")) 

plot.theme <- theme_cowplot() + theme(plot.title = element_text(hjust = 0.5, face = "plain"))

```

## remove genes function
```{r}
#remove TCR genes from VF lists
remove_genes <- function(features, species, tcr=TRUE, ig=TRUE, cell_cycle=TRUE, mito=TRUE, histone=TRUE, ribosome=TRUE) {
    print(paste("Before filtering:", length(features), "features"))

    if (species == "human") {
        if (tcr) {
            features <- features[!grepl("^TR[ABGD][VDJC]", features)]
        }
        if (ig) {
            features <- features[!grepl("^IG[HLK][VDJC]", features)]
        }
        if (cell_cycle) {
            features <- features[!(features %in% unlist(cc.genes))]
        }
        if (mito) {
            features <- features[!grepl("^MT-", features)]
        }
    }
    if (species == "mouse") {
        if (tcr) {
            features <- features[!grepl("^Tr[abgd][vdjc]", features)]
        }
        if (tcr) {
            features <- features[!grepl("^Tcr[abgd][vdjc]", features)]
        }
        if (ig) {
            features <- features[!grepl("^Ig[hlk][vdjc]", features)]
        }
        if (cell_cycle) {
            features <- features[!(features %in% stringr::str_to_title(unlist(cc.genes)))]
        }
        if (mito) {
            features <- features[!grepl("^Mt-", features)]
        }
      
       if (histone) {
            features <- features[!grepl("^Hist", features)]
       }
      
       if (ribosome) {
            features <- features[!grepl("^Rps", features)]
       }
      
      if (ribosome) {
            features <- features[!grepl("^Rpl", features)]
        }
      
    }

    print(paste("After filtering:", length(features), "features"))
    
    return(features)
}
```

## find variable features, and remove TCR, cell_cycle etc, genes - save as vf.clean
```{r}
# already ran find vf in integration workflow, but repeating to be safe

DefaultAssay(so) <- "integrated"
so <- FindVariableFeatures(so)
vf.clean <- remove_genes(VariableFeatures(so), 
                         species = "mouse" , tcr=T, ig=F, cell_cycle=T, mito=T, histone=T, ribosome=T)


```

## find DEGs
```{r}
DefaultAssay(so) <- "RNA"
all.markers <- FindAllMarkers(so, features = vf.clean, only.pos = T, logfc.threshold = 0.25, assay = "RNA", test.use = "wilcox", verbose = T)
```

## rename cell clusters - based on DEG 
```{r}
DimPlot(so, label = T)

# rename idents based on DEGs
so@meta.data$cell_cluster <- NA

so$cell_cluster[so$seurat_clusters == "0"] <-  "Tem_2"
so$cell_cluster[so$seurat_clusters == "1"] <-  "Tcm"
so$cell_cluster[so$seurat_clusters == "2"] <-  "Tem_1"
so$cell_cluster[so$seurat_clusters == "3"] <-  "Tex-Term"
so$cell_cluster[so$seurat_clusters == "4"] <-  "Tex-KLR"
so$cell_cluster[so$seurat_clusters == "5"] <-  "Tex-Int"
so$cell_cluster[so$seurat_clusters == "6"] <-  "Tex-Prog"
so$cell_cluster[so$seurat_clusters == "7"] <-  "Tex-ISG"
so$cell_cluster[so$seurat_clusters == "8"] <-  "Prolif"

Idents(so) <- factor(so$cell_cluster,
                           levels = c( "Tcm", "Tem_1", "Tem_2", "Tex-KLR", "Tex-Int", "Tex-Term", "Tex-Prog",  "Tex-ISG", "Prolif"))


DimPlot(so, cols = clusterpal) + #+ plot.theme.umap + ggtitle(label = "CD8 T cells") 
  geom_text(data = data.frame(x = 1, y = 2), x = 2.5, y = -5.3, label = paste("n =", length(so$barcode), "cells"), size = 4) + 
  plot.theme.umap +
  ggtitle("GP33+ T Cells")


so@meta.data %>% dplyr::count(cell_cluster, Group) %>% group_by(Group)%>% mutate(freq = n/sum(n)) %>% 
  filter(Group == "Cl13 d100+") %>% filter(grepl("Tex", cell_cluster)) %>% ungroup() %>% mutate(sum = sum(freq))
```

## plot DEG (all DEG)
```{r}
# fetch expression with dotplot
hm.df <- DotPlot(so, features = unique(all.markers$gene))$data %>% 
  select(c(avg.exp, features.plot, id)) %>% 
  pivot_wider(names_from = id, values_from = avg.exp) %>% 
  as.data.frame() %>%  `rownames<-`(.[,1]) %>% select(-features.plot) %>%
  t() %>% scale()

## annotations
mark.df <- data.frame('gene' = colnames(hm.df), 'n' = 1:length(colnames(hm.df))) 
genes <- c("Tox", "Pdcd1", "Tim3", "Havcr2", "Lag3", "Cd101", "Sell", "Il7r", "Ccr7", "Tcf7", "Myb", "Slamf6", 
           "Cx3cr1", "Zeb2", "Klrg1", "Xcl1", "Gzmb","Isg15", "Id3", "Mki67", "Cxcr6", "Ifng", "Pclaf", "Cdkn3", "Cd7", "Cd160", "Ifit3", "Ifi1","Ifi27l2a", "Cxcr3", "Lgals3", "S1pr5")

gene.annot <- columnAnnotation(anno = anno_mark(at = mark.df$n[mark.df$gene %in% genes], 
                                             labels = mark.df$gene[mark.df$gene %in% genes], 
                                             which = "row", padding = 0.8))

ha.clust <- rowAnnotation(`Cluster` = rownames(hm.df), 
                        col = list(`Cluster` = clusterpal),
                        show_legend = F)

# plot HM
hm.alldeg <- hm.df %>% 
  Heatmap(name = "Z-score", 
          left_annotation = ha.clust, top_annotation = gene.annot,
          col = viridis(n = 9, option = "magma") ,
          # col = zrev(brewer.pal(9, "RdBu")), 
          width = unit(5, "in"), height = unit(2.2, "in"), 
          show_column_dend = F, show_column_names = F, row_dend_width = unit(0.2, "cm"),
          cluster_rows = T, cluster_columns = T, 
          border = T, column_names_rot = 45
        )

hm.alldeg

```

## visualization of cluster distribution per group / sample
```{r}
# dim reduction plots per group 
umap.df <- data.frame(so@meta.data,  so@reductions$umap@cell.embeddings)

var.list <- c("Cl13 d21-26", "Cl13 d100+", "Arm" )  # vector of groups

plots <- lapply(var.list, function(x) {
  subset <- umap.df %>% subset(Group == x)
  umap <- ggplot(umap.df, aes(x = UMAP_1, y = UMAP_2)) +
    geom_point(color = "grey80", size = 0.3) +
    geom_point(data = subset, size = 0.3, aes(color = cell_cluster)) + 
    ggtitle(x) + 
    scale_color_manual(values = clusterpal)  + plot.theme.umap + NoLegend()

   umap %>% return()
   
})

do.call(plot_grid, c(plots, ncol = 3))

# Bar plots per each animal
so@meta.data %>% count(Group, REP, Sort, cell_cluster) %>% 
  group_by(Group, REP, Sort) %>% mutate(n = n/sum(n)) %>% 
  ggplot(aes(x = paste0(Group, Sort, REP), fill = cell_cluster, y = n)) + 
           geom_col() + 
  plot.theme + 
  scale_y_continuous(expand = expansion(mult = c(0.0, 0.02))) + 
  scale_fill_manual(values = clusterpal, name = "Cluster")  + 
  labs(x = "Sample", y = "Fraction of Cells") + rotate_x_text(90)


# plot out each sort on UMAP
var.list <- c("Arm", "Cl13 d21-26", "Cl13 d100+") # vector of groups

plots <- lapply(var.list, function(x) {
  subset <- umap.df %>% subset(Group == x) %>% rownames_to_column("Bar") %>% arrange(Bar)
 
  umap <- ggplot(umap.df, aes(x = UMAP_1, y = UMAP_2)) +
    geom_point(color = "grey90", size = 0.3) +
    geom_point(data = subset, size = 0.3, aes(color = Sort)) + 
    ggtitle(x) +
    plot.theme.umap 
  return(umap)
     
})

do.call(ggarrange, c(plots, ncol = 3, common.legend = T))

```
# export RDS
```{r}
saveRDS(so, file = paste0(path, "/data/1D_IntegratedSeuratObject_IDdClusters.rds"))

```


