# Now, digging a bit deeper into the DEG between groups. 
## the main question we're seeking to answer here is if chronic Tcm or Tem are more similar to their accute counterparts or Tex

# load common libraries + load SO + set paths
```{r}
suppressPackageStartupMessages({
  library(tidyverse) 
  library(Seurat)
  library(RColorBrewer)
  library(viridis)
  library(cowplot)
  library(ggridges)
  library(ggrepel)
  library(data.table)
  library(ComplexHeatmap)
  library(ggpubr)
  library(ComplexHeatmap)
  library(circlize)
  library(ggridges)
  library(writexl)
})

path <- "~/ChronicMemoryGithub_Submission/1_scRNATCRseqAtlas"
so <- readRDS(paste0(path, "/data/1D_IntegratedSeuratObject_IDdClusters.rds"))

```

# set plot aesthetics
```{r}
# palette
clusterpal <- c(
  "Tcm" = "#65c7ca",
  "Tem_1" = "#f8a523",
  "Tem_2" = "#f6d13a",
  "Tex-Term" = "#b31e0e",
  "Tex-Int" = "#de311e",
  "Tex-KLR" = "#E27E73",
  "Tex-ISG" = "#a0a1d1",
  "Tex-Prog" = "#a94d9a",
  "Prolif" = "#e6af7c",
  "Tem" = "#f6d13a"
  )


plot.theme.umap <- theme_cowplot() + theme(plot.title = element_text(hjust = 0.5), 
        axis.text = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5, face = "plain")) 

plot.theme <- theme_cowplot() + theme(plot.title = element_text(hjust = 0.5, face = "plain"))

plot.theme.box <- theme_bw() + 
  theme(plot.title = element_text(hjust =  0.5 , size = 14, color = "black"),
        plot.subtitle = element_text(hjust = 0.5 , size = 10, color = "black"), 
        axis.text = element_text(color = "black", size = 12),
        axis.title = element_text(size = 12),
        axis.ticks = element_line(color = "black", linewidth = 0.7), 
        strip.background = element_blank(), strip.text = element_text(hjust = 0, size = 11),
        panel.border = element_rect(colour = "black", linewidth = 0.7), 
        panel.grid = element_blank()
        )

```

# remove genes function
```{r}
#remove TCR genes from VF lists
remove_genes <- function(features, species, tcr=TRUE, ig=TRUE, cell_cycle=TRUE, mito=TRUE, histone=TRUE, ribosome=TRUE) {
    print(paste("Before filtering:", length(features), "features"))

    if (species == "human") {
        if (tcr) {
            features <- features[!grepl("^TR[ABGD][VDJC]", features)]
        }
        if (ig) {
            features <- features[!grepl("^IG[HLK][VDJC]", features)]
        }
        if (cell_cycle) {
            features <- features[!(features %in% unlist(cc.genes))]
        }
        if (mito) {
            features <- features[!grepl("^MT-", features)]
        }
    }
    if (species == "mouse") {
        if (tcr) {
            features <- features[!grepl("^Tr[abgd][vdjc]", features)]
        }
        if (tcr) {
            features <- features[!grepl("^Tcr[abgd][vdjc]", features)]
        }
        if (ig) {
            features <- features[!grepl("^Ig[hlk][vdjc]", features)]
        }
        if (cell_cycle) {
            features <- features[!(features %in% stringr::str_to_title(unlist(cc.genes)))]
        }
        if (mito) {
            features <- features[!grepl("^Mt-", features)]
        }
      
       if (histone) {
            features <- features[!grepl("^Hist", features)]
       }
      
       if (ribosome) {
            features <- features[!grepl("^Rps", features)]
       }
      
      if (ribosome) {
        }
      
    }

    print(paste("After filtering:", length(features), "features"))
    
    return(features)
}
```

# check seur obj and set cleaned variable features
```{r}
DimPlot(so)
DefaultAssay(so) <- "integrated"
VariableFeatures(so) %>% length()

vf.clean <- remove_genes(VariableFeatures(so), 
                         species = "mouse" , tcr=T, ig=F, cell_cycle=T, mito=T, histone=T, ribosome=T)
```

# add Phenotype / Group as an additional metadata column
```{r}
so$Group[so$Group == "Cl13 d21-26"] <- "Chronic Cl13"
so$Group[so$Group == "Cl13 d100+"] <- "Cleared Cl13"
so$Group[so$Group == "Arm"] <- "Arm"

# add new column for PhenoGroup
so@meta.data$Pheno_Group <- paste(sep = "__", so$Group, so$cell_cluster)

so@meta.data$Pheno_Group %>% table

```

# PC correlation
## to look at differences between clusters, we extract PC embeddings, then look at correlations of the top 30 PCs between groups
```{r}
library(Hmisc)

pca.df <- data.frame(so@meta.data %>% select(Group, Tet, Pheno_Group, cell_cluster), Embeddings(so, reduction = "pca")) %>% 
  filter(Pheno_Group %in%c("Arm__Tem_1", "Arm__Tem_2", "Arm__Tcm", 
                           "Chronic Cl13__Tex-Int" , "Chronic Cl13__Tex-KLR", "Chronic Cl13__Tex-Prog" , "Chronic Cl13__Tex-Term",
                           "Cleared Cl13__Tcm" ,  "Cleared Cl13__Tex-Prog", "Cleared Cl13__Tex-Term", "Cleared Cl13__Tem_1", "Cleared Cl13__Tem_2"
                          )) 

# group by PhenoGroup and summarized PCs
summarized.pcs <- pca.df %>% group_by(Pheno_Group) %>% 
  dplyr::summarise(across(where(is.numeric), median, na.rm = TRUE))


# find correlations across all groups
cors <- summarized.pcs %>% column_to_rownames("Pheno_Group") %>% t() %>% rcorr() 

# select only correlations of interest
r.values <- cors$r[1:7,8:12] 
p.values <- cors$P[1:7,8:12]

sig_matrix <- matrix(" " , nrow = nrow(r.values), ncol = ncol(r.values))
colnames(sig_matrix) <- colnames(r.values)
rownames(sig_matrix) <- rownames(r.values)

# correct p values with binferoni 
sig_matrix[p.values <=  0.05 / p.values %>% length()] <- "*" 
sig_matrix[p.values <=  0.01 / p.values %>% length()] <- "**" 
sig_matrix[p.values <=  0.001 / p.values %>% length()] <- "***"
sig_matrix <- t(sig_matrix)

# rename row and column names
rownames(r.values) <- gsub(".*__", "", rownames(r.values))
colnames(r.values) <- gsub(".*__", "", colnames(r.values))


# plot
row.annot <- rowAnnotation(` `  = colnames(r.values), col = list(` ` = clusterpal), show_legend = F)
col.annot <- columnAnnotation(` `  = rownames(r.values), col = list(` ` = clusterpal), show_legend = F)
split.vector <-  ifelse(grepl("Arm", rownames(p.values)), "Arm", "Chronic Cl13")

hm.cor <- r.values %>% t %>% Heatmap(
                      col = viridis(9) , border = T, 
                      width = unit(7*0.8, "cm"), height = unit(5*0.8, "cm"), name = "Pearson R", 
                      cluster_rows = T, cluster_columns = T, show_column_dend = F, row_dend_width = unit(0.2, "cm"), row_dend_side = "right",
                      row_title = "Cleared Cl13 Clusters" , 
                      row_title_side = "left", column_title_side = "bottom", row_names_side = "left",
                      left_annotation = row.annot, bottom_annotation = col.annot,
                      cell_fun = function(j, i, x, y, width, height, fill) {grid.text(sig_matrix[i, j], x, y, gp = gpar(fontsize = 12, col = "white"))},
                      column_split = split.vector
                  )

hm.cor

```

# deg between cTcm or cTem with other groups
```{r}
# group Tem_1 and Tem_2 together for simplicity
so$Pheno_Group2 <- gsub("_\\d+$", "", so$Pheno_Group)
test.list <- c("Arm__Tem", "Arm__Tcm","Chronic Cl13__Tex-Int" , "Chronic Cl13__Tex-KLR", "Chronic Cl13__Tex-Prog" , "Chronic Cl13__Tex-Term")

# Calcualte DEG vs cTcm ================================================================

deg.tcm <- lapply(test.list, function(x){

    ## vector of up cells and down cells
    cells.up <- subset(so, Pheno_Group2 == "Cleared Cl13__Tcm") %>% colnames()
    cells.down <- subset(so, Pheno_Group2 == x) %>% colnames()
    
    deg <- FindMarkers(object = so, 
                       ident.1 = cells.up, ident.2 = cells.down, 
                       logfc.threshold = 0, assay = "RNA", features = vf.clean)
    
    deg <- deg %>% mutate(gene = rownames(deg)) %>%
      mutate(sig = dplyr::if_else(p_val_adj < 0.01 & avg_log2FC > 0.25, "up", "ns")) %>% 
      mutate(sig = dplyr::if_else(p_val_adj < 0.01 & avg_log2FC < -0.25,  "down", sig)) %>%
      filter(sig != "ns") %>% 
      mutate(cluster = x)
    
    return(deg)

})

# Calcualte DEG vs cTem ================================================================

deg.tem <- lapply(test.list, function(x){

    ## vector of up cells and down cells
    cells.up <- subset(so, Pheno_Group2 == "Cleared Cl13__Tem") %>% colnames()
    cells.down <- subset(so, Pheno_Group2 == x) %>% colnames()
    
    deg <- FindMarkers(object = so, 
                       ident.1 = cells.up, ident.2 = cells.down, 
                       logfc.threshold = 0, assay = "RNA", features = vf.clean)
    
    deg <- deg %>% mutate(gene = rownames(deg)) %>%
      mutate(sig = dplyr::if_else(p_val_adj < 0.01 & avg_log2FC > 0.25, "up", "ns")) %>% 
      mutate(sig = dplyr::if_else(p_val_adj < 0.01 & avg_log2FC < -0.25,  "down", sig)) %>%
      filter(sig != "ns") %>% 
      mutate(cluster = x)
    
    return(deg)

})

# PLOT # of Dif expr. genes ==============================================

plot.df <- do.call(rbind, deg.tcm) %>% separate(cluster, into = c("Group", "Cluster"), sep = "__") %>% 
  count(sig, Group, Cluster) 
sum.df <- do.call(rbind, deg.tcm) %>% separate(cluster, into = c("Group", "Cluster"), sep = "__") %>% count(Group, Cluster) 


p1 <- plot.df %>%
  ggplot(aes(x = fct_reorder(paste(Group, Cluster), n), y = n)) + 
  geom_col(aes(fill = Cluster, alpha = sig), show.legend = F) + 
  geom_text(data = sum.df, aes(label = n, y = n+34)) +
  plot.theme + 
  scale_fill_manual(values = clusterpal) + scale_alpha_manual(values = c(0.6,1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) + 
  labs(title = "DEG vs cTcm", x = element_blank(), y = "# of DEG") + 
  rotate_x_text(45)

plot.df <- do.call(rbind, deg.tem) %>% separate(cluster, into = c("Group", "Cluster"), sep = "__") %>% 
  count(sig, Group, Cluster) 
sum.df <- do.call(rbind, deg.tem) %>% separate(cluster, into = c("Group", "Cluster"), sep = "__") %>% count(Group, Cluster) 

p2 <-  plot.df %>%
  ggplot(aes(x = fct_reorder(paste(Group, Cluster), n), y = n)) + 
  geom_col(aes(fill = Cluster, alpha = sig), show.legend = F) + 
  geom_text(data = sum.df, aes(label = n, y = n+30)) +
  plot.theme + 
  scale_fill_manual(values = clusterpal) + scale_alpha_manual(values = c(0.6,1)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) + 
  labs(title = "DEG vs cTem", x = element_blank(), y = "# of DEG") + 
  rotate_x_text(45)

plot_grid(p1,p2)


```


# save tables 
```{r}

tcm.genes <- do.call(rbind, deg.tcm) 
tem.genes <- do.call(rbind, deg.tem)  

# tcm.genes %>% filter(sig != "ns") %>% select(gene, avg_log2FC, p_val, p_val_adj, sig, cluster) %>% write_xlsx(paste0(outs, "/2C_DEG_Tcm.xlsx"))
# tem.genes %>% filter(sig != "ns") %>% select(gene, avg_log2FC, p_val, p_val_adj, sig, cluster) %>% write_xlsx(paste0(outs, "/2C_DEG_Tem.xlsx"))
```

# add common exhaustion module and plot per group
```{r}
# common exh. score
## use the total score from Daniel Et al Nat Immuno
common.ex.genes <- read.csv("~/Resources/LCMV atlas tables/LCMV_common_ex_score.csv")$Common_Ex_Module
common.ex.genes <- common.ex.genes[common.ex.genes %in% vf.clean] %>% list()

DefaultAssay(so) <- "RNA"
so <- AddModuleScore(so, features = common.ex.genes, name = "CommonExMod", assay = "RNA")

VlnPlot(so, features = "CommonExMod1",  cols = clusterpal, pt.size = 0, sort = "decreasing") & plot.theme & rotate_x_text(45) & NoLegend()
FeaturePlot(so, features = "CommonExMod1",  cols = viridis(9, option = "magma")) & plot.theme.umap & ggtitle("Exhaustion Module")
# ggsave("2C - CommonExhModule.png", path = outs, width = 4, height = 3.5)
# ggsave("2C - CommonExhModule.pdf", path = outs, width = 4, height = 3.5)

# vln split by group
so@meta.data %>% 
  filter(Pheno_Group %in%c("Arm__Tcm", "Arm__Tem_1", "Arm__Tem_2",
                           "Chronic Cl13__Tex-Int" , "Chronic Cl13__Tex-KLR", "Chronic Cl13__Tex-Prog" , "Chronic Cl13__Tex-Term",
                           "Cleared Cl13__Tcm" ,  "Cleared Cl13__Tex-Prog", "Cleared Cl13__Tex-Term", "Cleared Cl13__Tex-Int", "Cleared Cl13__Tex-KLR",  "Cleared Cl13__Tem_1", "Cleared Cl13__Tem_2"
                          )) %>%
  mutate(Group = factor(Group, levels = c("Chronic Cl13", "Cleared Cl13", "Arm"))) %>%
  ggplot(aes(x = fct_reorder(cell_cluster, CommonExMod1) , y = CommonExMod1, fill = cell_cluster)) + 
  geom_violin(scale = "width", draw_quantiles = 0.5, show.legend = F, color = "black") + 
  ggforce::facet_row(vars(Group), scales = 'free', space = 'free') +
  plot.theme + theme(strip.background = element_blank()) + rotate_x_text(45) + 
  labs(x = element_blank(), y = "Exhaustion Module") +
  scale_fill_manual(values = clusterpal) +
  scale_y_continuous(limits = c(-0.6, 1.5))

# ggsave("2B - CommonExhModule.pdf", path = outs, width = 7, height = 2.8)

# median per sample
so@meta.data %>% 
  filter(Pheno_Group %in%c("Arm__Tcm", "Arm__Tem_1", "Arm__Tem_2",
                           "Chronic Cl13__Tex-Int" , "Chronic Cl13__Tex-KLR", "Chronic Cl13__Tex-Prog" , "Chronic Cl13__Tex-Term",
                           "Cleared Cl13__Tcm" ,  "Cleared Cl13__Tex-Prog", "Cleared Cl13__Tex-Term", "Cleared Cl13__Tex-Int", "Cleared Cl13__Tex-KLR",  "Cleared Cl13__Tem_1", "Cleared Cl13__Tem_2"
                          )) %>%
  mutate(Group = factor(Group, levels = c("Chronic Cl13", "Cleared Cl13", "Arm"))) %>% 
  group_by(Pheno_Group) %>% summarise(CommonExMod1 = median(CommonExMod1))



 
```


