# here we will dive into clonal behaviors of T cells in during Cl13 infection - either during the chronic phase or after viral clearance

# load common libraries + load SO + set plot themes
```{r}
suppressPackageStartupMessages({
  library(tidyverse) 
  library(Seurat)
  library(RColorBrewer)
  library(viridis)
  library(cowplot)
  library(ggridges)
  library(ggrepel)
  library(data.table)
  library(ComplexHeatmap)
  library(ggpubr)
  library(ComplexHeatmap)
  library(circlize)
  library(ggridges)
})

# read in seurat object
path <- "~/ChronicMemoryGithub_Submission/1_scRNATCRseqAtlas"
so <- readRDS(paste0(path, "/data/1D_IntegratedSeuratObject_IDdClusters.rds"))

# add a clone_id column to include cdr3, replicate, and group - there shouldnt be any overlapping TCRs but just to be safe
so@meta.data$clone_id <- paste(so$Group, so$orig.ident,so$REP ,  so$cdr3s_nt, sep = "_")
```

# set plot aesthetics
```{r}
# palette
clusterpal <- c(
  "Tcm" = "#65c7ca",
  "Tem_1" = "#f8a523",
  "Tem_2" = "#f6d13a",
  "Tex-Term" = "#b31e0e",
  "Tex-Int" = "#de311e",
  "Tex-KLR" = "#E27E73",
  "Tex-ISG" = "#a0a1d1",
  "Tex-Prog" = "#a94d9a",
  "Prolif" = "#e6af7c"
  )

grouppal <- c("Chronic Cl13" = "#941A1D", "Cleared Cl13" = "#F37C79", "Arm" = "#498DCB")

plot.theme.umap <- theme_cowplot() + theme(plot.title = element_text(hjust = 0.5), 
        #panel.border = element_rect(colour = "black", fill=NA, linewidth = 0.5, color = "grey80"),
        axis.text = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5, face = "plain")) 

plot.theme <- theme_cowplot() + theme(plot.title = element_text(hjust = 0.5, face = "plain"))

plot.theme.box <- theme_bw() + 
  theme(plot.title = element_text(hjust =  0.5 , size = 14, color = "black"),
        plot.subtitle = element_text(hjust = 0.5 , size = 10, color = "black"), 
        axis.text = element_text(color = "black", size = 12),
        axis.title = element_text(size = 12),
        axis.ticks = element_line(color = "black", linewidth = 0.7), 
        strip.background = element_blank(), strip.text = element_text(hjust = 0, size = 11),
        panel.border = element_rect(colour = "black", linewidth = 0.7), 
        panel.grid = element_blank()
        )
```

# remove genes function
```{r}
#remove TCR genes from VF lists
remove_genes <- function(features, species, tcr=TRUE, ig=TRUE, cell_cycle=TRUE, mito=TRUE, histone=TRUE, ribosome=TRUE) {
    print(paste("Before filtering:", length(features), "features"))

    if (species == "human") {
        if (tcr) {
            features <- features[!grepl("^TR[ABGD][VDJC]", features)]
        }
        if (ig) {
            features <- features[!grepl("^IG[HLK][VDJC]", features)]
        }
        if (cell_cycle) {
            features <- features[!(features %in% unlist(cc.genes))]
        }
        if (mito) {
            features <- features[!grepl("^MT-", features)]
        }
    }
    if (species == "mouse") {
        if (tcr) {
            features <- features[!grepl("^Tr[abgd][vdjc]", features)]
        }
        if (tcr) {
            features <- features[!grepl("^Tcr[abgd][vdjc]", features)]
        }
        if (ig) {
            features <- features[!grepl("^Ig[hlk][vdjc]", features)]
        }
        if (cell_cycle) {
            features <- features[!(features %in% stringr::str_to_title(unlist(cc.genes)))]
        }
        if (mito) {
            features <- features[!grepl("^Mt-", features)]
        }
      
       if (histone) {
            features <- features[!grepl("^Hist", features)]
       }
      
       if (ribosome) {
            features <- features[!grepl("^Rps", features)]
       }
      
      if (ribosome) {
        }
      
    }

    print(paste("After filtering:", length(features), "features"))
    
    return(features)
}
```

# Clonal Diversity and # of clones per sample
### we found that there's less clonal diversity in cleared cl13 than either chronic Cl13 or Acute - it was an interesting finding and tracks with what we think that only select Tex clones can persist long term, but didnt focus on this in the mansucript
```{r}
suppressPackageStartupMessages({
library(ineq)
library(vegan)
})

# create a dataframe(diversity.df) of clone counts per sample
diversity.df <-so@meta.data %>% filter(chains =="ab") %>% 
  dplyr::count(clone_id, Group, REP, orig.ident) %>% filter(n > 1) %>% 
  group_by(orig.ident, Group, REP) %>% mutate(freq = n / sum(n)) %>% 
  ungroup() %>%
  mutate(SampleID = paste(orig.ident, Group, REP, sep = "_")) %>% 
  mutate(Group = ifelse(Group == "Cl13 d100+", "Cleared Cl13", Group)) %>% 
  mutate(Group = ifelse(Group == "Cl13 d21-26", "Chronic Cl13", Group))

# split tcr_clean into tcr.list by samples
diversity.list <- split(diversity.df, diversity.df$SampleID)

# calculate gini index for each sample separately  ===========================
gini.list <- lapply(diversity.list, function(x) {
 gini <- Gini(x$freq)
 df <- x %>% group_by(SampleID) %>% summarise( 
    across(c(Group, REP, orig.ident), first))
 df$Gini <- gini
 return(df)
})

gini.results <- do.call(rbind, gini.list)

p1 <- gini.results %>% ggplot(aes(x = fct_reorder(Group, Gini), y = Gini)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(width = 0.23, aes(color = Group), show.legend = F, size = 2) +
  stat_compare_means(comparisons = list(c("Cleared Cl13", "Chronic Cl13"), c("Arm", "Cleared Cl13"), c("Arm", "Chronic Cl13")), method = "t.test", tip.length = 0) + 
  scale_color_manual(values = grouppal) + 
  plot.theme + rotate_x_text(45) + labs(x = element_blank(), y = "Gini Index")  + 
  scale_y_continuous(expand = expansion(mult = 0.15))

#  calculate shannon diversity for each sample separately ===========================
shannon.list <- lapply(diversity.list, function(x) {
 div <- diversity(x$freq,  index = "shannon")
 df <- x %>% group_by(SampleID) %>% summarise(
    across(c(Group, REP, orig.ident), first))
 df$Shannon <- div
 return(df)
})

shannon.results <- do.call(rbind, shannon.list)

p2 <- shannon.results %>% ggplot(aes(x = fct_reorder(Group, -1*Shannon), y = Shannon)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(width = 0.23, aes(color = Group), show.legend = F, size = 2) +
  stat_compare_means(comparisons = list(c("Cleared Cl13", "Chronic Cl13"), c("Arm", "Cleared Cl13"), c("Arm", "Chronic Cl13")), method = "t.test", tip.length = 0) + 
  scale_color_manual(values = grouppal) + 
  plot.theme + rotate_x_text(45) + labs(x = element_blank(), y = "Shannon Diversity") + 
  scale_y_continuous(expand = expansion(mult = 0.15))

# add # of expanded clones per mouse  ===========================

p3 <- diversity.df %>% count(Group, SampleID) %>% ggplot(aes(x = Group, y = n)) +geom_boxplot(outlier.shape = NA) + 
  geom_jitter(width = 0.23, aes(color = Group), show.legend = F, size = 2) +
  # stat_compare_means(comparisons = list(c("Cleared Cl13", "Chronic Cl13"), c("Arm", "Cleared Cl13"), c("Arm", "Chronic Cl13")), method = "t.test", tip.length = 0) + 
  scale_color_manual(values = grouppal) + 
  plot.theme + rotate_x_text(45) + labs(x = element_blank(), y = "Number of Clones per Mouse")  + 
  scale_y_continuous(limits = c(0,100))

plot_grid(p3, p2, p1, ncol = 3)

# ggsave(path = outs, filename = "3C - DiversityIndecies_plusCloneNumebr.pdf", width = 7.5, height = 3.5)

```

# Cl13 d100+: define clone behaviors by kmeans clustering of intraclonal proportions
```{r}
# find TCRab clones that are expanded (clone size>1)
exp.clones <- so@meta.data %>% filter(Group == "Cl13 d100+", chains =="ab") %>% 
  dplyr::count(clone_id, name = "CloneSize") %>% 
  mutate(CloneFreq = CloneSize / sum(CloneSize)) %>% ungroup %>% 
  select(clone_id, CloneFreq, CloneSize) %>% filter(CloneSize > 1)

# determine phenotype proportion per clone
clone.df <- so@meta.data %>% filter(clone_id %in% exp.clones$clone_id) %>% 
  dplyr::count(cell_cluster, clone_id) %>%
  group_by(clone_id) %>% mutate(n = n/sum(n)) %>%
  pivot_wider(names_from = cell_cluster, values_from = n) %>%
  ungroup() %>% as.data.frame() %>%
  mutate(across(everything(), ~replace_na(.x, 0))) %>% 
  left_join(exp.clones) 

head(clone.df)

# cluster clones based on clone proportions
set.seed(22)
clusters <- kmeans(clone.df %>% select(-c(clone_id, CloneSize, CloneFreq)), centers = 5)
clone.df$Clone_Behavior <- clusters$cluster

# rename clusters based on phenotype prop
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "1"] <- "Tem_2 Bias"
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "2"] <- "Tem_1 Bias"
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "3"] <- "Divergent"
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "4"] <- "Tcm Bias"
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "5"] <- "Tex Bias"

# heatmaps
col.annot <- columnAnnotation(` `  = select(clone.df, -c(clone_id, Clone_Behavior, CloneFreq,  CloneSize)) %>% colnames(), 
              col = list(` ` = clusterpal), show_legend = F)
size.annot <- rowAnnotation(` ` = anno_barplot(clone.df$CloneFreq %>% log2, baseline = "min"))

clone.hm <- clone.df %>% select(-c(clone_id, Clone_Behavior, CloneSize, CloneFreq)) %>% 
  Heatmap(name = "Fraction of Clone",
          show_column_names = T, border = T, col= brewer.pal(9, "Purples"), 
          row_split = clone.df$Clone_Behavior,
          show_row_dend = F, show_column_dend = T,
          row_title_rot = 0, 
          width = unit(9*0.7, "cm"), height = unit(9*0.7, "cm"), column_dend_height = unit(0.2, "cm"),
          top_annotation = col.annot, right_annotation = size.annot,
          column_title = paste0("Cleared Cl13 n=", clone.df$clone_id %>% length(), " Clones >1 Cell") 
          )
clone.hm

#save for future use
clone.df.Cl13Cleared <- clone.df
```


# Cl13 Chronic: define clone behaviors by the same method
```{r}

# find TCRab clones that are expanded (clone size>1)
exp.clones <- so@meta.data %>% filter(Group == "Cl13 d21-26", chains =="ab") %>% 
  dplyr::count(clone_id, name = "CloneSize") %>% 
  mutate(CloneFreq = CloneSize / sum(CloneSize)) %>% ungroup %>% 
  select(clone_id, CloneFreq, CloneSize) %>% filter(CloneSize > 1)

# phenotype proportion per clone
clone.df <- so@meta.data %>% filter(clone_id %in% exp.clones$clone_id) %>% 
  dplyr::count(cell_cluster, clone_id) %>%
  group_by(clone_id) %>% mutate(n = n/sum(n)) %>%
  pivot_wider(names_from = cell_cluster, values_from = n) %>%
  ungroup() %>% as.data.frame() %>%
  mutate(across(everything(), ~replace_na(.x, 0))) %>% 
  left_join(exp.clones) 

head(clone.df)


# cluster clones based on clone proportions
set.seed(22)
clusters <- kmeans(clone.df %>% select(-c(clone_id, CloneSize, CloneFreq)), centers = 5)
clone.df$Clone_Behavior <- clusters$cluster

# rename clusters
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "1"] <- "Tex-KLR Bias"
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "2"] <- "Tex-Term Bias"
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "3"] <- "Divergent"
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "4"] <- "Tex-Prog Bias"
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "5"] <- "Tex-Int Bias"


# heatmaps
col.annot <- columnAnnotation(` `  = select(clone.df, -c(clone_id, Clone_Behavior, CloneFreq,  CloneSize)) %>% colnames(), 
              col = list(` ` = clusterpal), show_legend = F)
size.annot <- rowAnnotation(` ` = anno_barplot(clone.df$CloneFreq %>% log2, baseline = "min"))

clone.hm <- clone.df %>% select(-c(clone_id, Clone_Behavior, CloneSize, CloneFreq)) %>% 
  Heatmap(name = "Fraction of Clone",
          show_column_names = T, border = T, col= brewer.pal(9, "Purples"), 
          row_split = clone.df$Clone_Behavior,
          show_row_dend = F, show_column_dend = T,
          row_title_rot = 0, 
          width = unit(9*0.7, "cm"), height = unit(9*0.7, "cm"), column_dend_height = unit(0.2, "cm"),
          top_annotation = col.annot, right_annotation = size.annot,
          column_title = paste0("Chronic Cl13 n=", clone.df$clone_id %>% length(), " Clones >1 Cell") 
          )

clone.hm

clone.df.cl13chronic <- clone.df
```

# Cl13 day 100+ : plot Top Clones UMAP 
```{r}
# select some of the top clones from each behavior 
clone.df.Cl13Cleared %>% group_by(Clone_Behavior) %>% slice_max(n = 20, order_by = CloneSize)# %>% select(CloneSize, Clone_Behavior, clone_id)

# dim recution plots
umap.df <- data.frame(so@meta.data,  so@reductions$umap@cell.embeddings)

# vector of clones
var.list <- c("Cl13 d100+_CR119_3_TRB:TGTGCCAGCTCTCTCCATCGCTCCGACTACACCTTC;TRA:TGTGCTATGGAACTTGACTATGCAAACAAGATGATCTTT",  #Tcm
              "Cl13 d100+_CR119_5_TRB:TGTGCCAGCAGCTTCTCCTATGAACAGTACTTC;TRA:TGTGCTACCCATGACACAAATGCTTACAAAGTCATCTTT", # Tem-1
              "Cl13 d100+_CR119_4_TRB:TGTGCCAGCAGTGACTGGACAGACACAGAAGTCTTCTTT;TRA:TGTGCAGTGAGGGGCAATAACAGAATCTTCTTT", # Tem-2
              "Cl13 d100+_CR119_1_TRB:TGTGCTAGCAGTTTAATACAATCAGAAGTCTTCTTT;TRA:TGCGCAGTCAGTGAAGGTTCAGCCTTAGGGAGGCTGCATTTT", # Tex
              "Cl13 d100+_CR119_3_TRB:TGTGCCAGCAGCCCCGACTGGGGGGAAGACACCCAGTACTTT;TRA:TGTGCTATGAGAGAGTGGGCATCCTCCTCCTTCAGCAAGCTGGTGTTT" # divergent
              ) 

plots <- lapply(var.list, function(x) {
  subset <- umap.df %>% subset(clone_id == x)
  umap <- ggplot(umap.df, aes(x = UMAP_1, y = UMAP_2)) +
    geom_point(color = "grey80", size = 1) +
    geom_point(data = subset, size = 1, aes(color = cell_cluster)) + 
    scale_color_manual(values = clusterpal)  + plot.theme.umap + NoLegend()

   umap %>% return()
   
})

do.call(plot_grid, c(plots, ncol = 5))

```

# Cl13 day chronic : plot Top Clones on UMAP 
```{r}
# select some of the top clones from each behavior 
clone.df.cl13chronic %>% group_by(Clone_Behavior) %>% slice_max(n = 5, order_by = CloneSize)# %>% select(CloneSize, Clone_Behavior, clone_id)

# dim recution plots
umap.df <- data.frame(so@meta.data,  so@reductions$umap@cell.embeddings)

# vector of clones
var.list <- c("Cl13 d21-26_CR119_5_TRB:TGTGCTAGCAGTTTAGTGAGTCAAAACACCTTGTACTTT;TRA:TGTGCTATGTCTTCTGGCAGCTGGCAACTCATCTTT", #term
              "Cl13 d21-26_CR82_2_TRB:TGTGCCAGCAGCTTATGGGGGGGTGCAGAAACGCTGTATTTT;TRA:TGTGCTACAGATACGCGATACAACGTGCTTTACTTC", # int
              "Cl13 d21-26_CR119_3_TRB:TGTGCCTGGAGGGACTGGGAGAACACCGGGCAGCTCTACTTT;TRA:TGTGCAGCAAGCCCCTCGAACACGGGTTACCAGAACTTCTATTTT", # klr
              "Cl13 d21-26_CR119_1_TRB:TGTGCCAGCAGCTCCGGGTGGGAACAGTACTTC;TRA:TGTGCTACCCATGACACAAATGCTTACAAAGTCATCTTT", # prog
              "Cl13 d21-26_CR119_1_TRB:TGTGCCAGCAGTATAAAGGGGGGCTCCTATGCTGAGCAGTTCTTC;TRA:TGTGCTATGACCAACTATGGAAATGAGAAAATAACTTTT"  #Divergent
              ) 

plots <- lapply(var.list, function(x) {
  subset <- umap.df %>% subset(clone_id == x)
  umap <- ggplot(umap.df, aes(x = UMAP_1, y = UMAP_2)) +
    geom_point(color = "grey80", size = 1) +
    geom_point(data = subset, size = 1, aes(color = cell_cluster)) + 
    scale_color_manual(values = clusterpal)  + plot.theme.umap + NoLegend()

   umap %>% return()
   
})

do.call(plot_grid, c(plots, ncol = 5))

```


# clone sizes of different behaviors
```{r}
clone.pal <- c("Tcm / Tem_1 Bias" = "#5CB865", 
               "Tcm Bias" = "#65c7ca",
               "Tem_1 Bias" = "#f8a523",
               "Tem_2 Bias" = "#f6d13a",
               "Divergent" = "#e6af7c",
               "Tex-Term Bias" = "#b31e0e",
               "Tex-Prog Bias" = "#a94d9a",
               "Tex-KLR Bias" = "#E27E73",
               "Tex-Int Bias"  = "#de311e",
               "Tex Bias" = "#de311e"
               )

clone.df <- rbind(
  select(clone.df.cl13chronic, clone_id, CloneSize, CloneFreq, Clone_Behavior),
  select(clone.df.Cl13Cleared, clone_id, CloneSize, CloneFreq, Clone_Behavior)) %>% 
  separate("clone_id", into = c("Group", "Experiment", "Rep", "cdr3s_nt"), sep = "_")
  
clone.df$Group%>% table


clone.df %>% ggplot(aes(x = fct_reorder(paste(Group, Clone_Behavior,sep = "---"), CloneFreq), y = log2(CloneFreq))) + 
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, aes(color = Clone_Behavior), show.legend = F) + 
  # facet_grid(~Group) +
  ggforce::facet_row("Group", scales = "free") +
  plot.theme + theme(strip.background = element_blank()) +  
  scale_x_discrete(labels = function(x) gsub(".*---", "", x)) + 
  scale_color_manual(values = clone.pal) + coord_flip() +
  labs(x = element_blank(), y = "Log2 Clone Freq")
  


```




