# in this experiment we longtidunally track clones from the blood at 21 dpi to 100+ dpi in the spleen
## this data was generated alongside anti-PDL1 clone tracing data (section 9), but we're focusing only on the control (PBS treated) group for this experiment

## load libraries for
```{r}
suppressPackageStartupMessages({
  library(tidyverse) 
  library(ComplexHeatmap)
  library(cowplot)
  library(ggpubr)
  library(readxl)
  library(RColorBrewer)
  library(MetBrewer)
  library(ggsankey)
  library(viridis)
  library(Hmisc)
  library(emmeans)
})
```

## set directories and plot aesthetics
```{r}
path <- "/Users/craposo/Desktop/Github Upload/5_LongitudinalTCRseq"


plot.theme <- theme_cowplot() + theme(plot.title = element_text(hjust = 0.5, face = "plain"), 
                                      strip.background = element_rect(fill = "transparent"), strip.text = element_text(size = 14)
                                      )

pop.pal <-  c(
  "Tcm" = "#65c7ca", 
  "Tex-KLR" = "#E27E73" ,
  "Tex-Int" = "#de311e", 
  "Tex-Term" = "#b31e0e", 
  "Tex-Prog" = "#a94d9a",
  "Tem_1" = "#f8a523",
  "Tem_2" = "#f6d13a", 
  
  "d100_Tcm" = "#65c7ca", 
  "d100_Tex-Term" = "#b31e0e", 
  "d100_Tex-Prog" = "#a94d9a",
  "d100_Tem_1" = "#f8a523",
  "d100_Tem_2" = "#f6d13a", 
  
  "d021_Tex-Int" = "#de311e", 
  "d021_Tex-KLR" = "#E27E73", 
  "d021_Tex-Prog" = "#a94d9a"
  

  )

cluster.pal <- c(
  "Divergent" = "#e6af7c", 
  "Tcm Bias" = "#65c7ca",
  "Tem_1 Bias" = "#f8a523",
  "Tem_2 Bias" = "#f6d13a", 
  "Tex-Prog Bias" = "#a94d9a", 
  "Tex-Term Bias" = "#b31e0e", 
  "Tex-KLR Bias" = "#E27E73" ,
  "Tex-Int Bias" = "#de311e"
)


```

# import bTCR-seq data
```{r}
# Specify the directory path and get all files in the directory
bTCR_datapath <- paste0(path, "/data/rawdata")
bTCR_datapath

samples <- list.files(path = bTCR_datapath)


# read in all csvs and append sample ID
bTCR_raw_list <- lapply(samples, function(x){
  sample.id <- substr(x, 27, 32)
  path <- paste0(bTCR_datapath, "/TCRlongtiduinal_Processed_" , sample.id , "_pep.csv")
  data <- read.csv(path)
  data$Sample.id <- sample.id
  return(data)
}) 
  

# append all TCR raw data together
bTCR_rawdata <-  do.call(rbind, bTCR_raw_list)

# read in sample IDs and append to bTCR_rawdata
sample.id <- read.csv(paste0(path, "/data/SampleInfo.csv")) %>% select(Sample, Sample.id)

bTCR_rawdata <- merge(bTCR_rawdata, sample.id, by = "Sample.id") %>% 
  mutate(CDR3.nuc.= toupper(CDR3.nuc.)) # make CDR3 all upper case to make futures things easier

# select only the samples that we want for analysis
bTCR_rawdata <- test <- bTCR_rawdata %>% filter(grepl("Ctrl", Sample), !grepl("d35", Sample))

bTCR_rawdata$Sample %>% table
bTCR_rawdata %>% head()

# select only the cdr3 and count columns
cdr3.df <- bTCR_rawdata %>% select(CDR3.nuc. , copy, Sample) %>%
  mutate(Rep = ifelse(substr(Sample , nchar(Sample) -1, nchar(Sample)) == "_r", "R2", "R1")) %>%  # add replicate column
  mutate(Sample = ifelse(Rep == "R2", substr(Sample, 1, nchar(Sample) -2), Sample))  # remove replicate from sample ID
```

# check qc and average the bTCR-seq data across technical replicates
```{r}
# QC check on replicate samples
## everything looks highly concordant between replicates
cdr3.df %>% pivot_wider(id_cols = c(CDR3.nuc. , Sample), values_from = copy, names_from = Rep, values_fn = {sum}) %>%
  mutate(across(everything(), ~ ifelse(is.na(.), 0, .))) %>%
  ggplot(aes(x = R1, y = R2)) + 
    geom_point(alpha = .5, size = 0.5) + 
    geom_smooth(method = "lm", color = "red", fill = "transparent") +
    facet_wrap(~Sample, scales = "free") + 
    stat_cor() + scale_x_log10() + scale_y_log10() + theme_bw() + ggtitle("all clones")

# average clone clounts across replicates
## make vector only where clone is in both replicates
both.rep.clones <- cdr3.df %>% pivot_wider(id_cols = c(CDR3.nuc. , Sample), values_from = copy, names_from = Rep, values_fn = {sum}) %>%
  mutate(across(everything(), ~ ifelse(is.na(.), 0, .))) %>% 
  filter(R1 >0, R2 > 0) %>% 
  select(CDR3.nuc.) %>% c()

## create new clone.df with average reads per sample
clone.df <- cdr3.df %>% filter(CDR3.nuc. %in% both.rep.clones$CDR3.nuc.) %>% # remove postive control, select only shared clones
  group_by(Rep, Sample, CDR3.nuc.) %>% mutate(ReadFrac = sum (copy)) %>% ungroup() %>% # any TCRs with the same CDR3 add together
  group_by(Rep, Sample) %>% mutate(ReadFrac = copy / sum (copy)) %>% # convert copies to fraction of reads
  ungroup() %>% group_by(Sample, CDR3.nuc.) %>% select(-c(Rep, copy)) %>% summarise(ReadFrac = mean(ReadFrac)) %>% # average read fraction between the two reads
  dplyr::rename(TRB_CDR3 = "CDR3.nuc.")

# add in cell numbers from sort - adjust the read fractions to cell # of population 
sampleinfo <- read_xlsx(paste0(path, "/data/CellNumbersSorted.xlsx")) %>%
  mutate(Sample = str_replace_all(Sample, " ", "_")) %>% select("Sample", "ActualCellNumbner_FlowData", "SortedCellNumber") %>%
  dplyr::rename(TotalCellNumber = "ActualCellNumbner_FlowData")

clone.df <- merge(clone.df, sampleinfo, by = "Sample") %>% 
  mutate(Count = ReadFrac * TotalCellNumber) %>% # count is based on total # of cells in population (based on counting beads)
  mutate(Count_Sort = ReadFrac * SortedCellNumber) %>% # count is based on total # of cells sorted from population (based on sorter #)
  ungroup()


# mutate sample ID into more useful metadata and rename the samples
## add total clone size and group into TCR
clone.df.pheno <- clone.df %>% 
  separate(Sample, into = c('experiment', 'dpi', 'group', 'sort')) %>% 
  group_by(TRB_CDR3, dpi, group) %>% mutate(CloneSize = sum(Count)) %>% mutate(CloneSize_Sort = sum(Count_Sort))%>% ungroup() %>% # 
  mutate(sort = ifelse(sort == 'cTcm', 'Tcm', paste0("Tex-", sort)), 
          sort = ifelse(sort == 'Tex-Int' & dpi == "d100", 'Tem_2', sort),
          sort = ifelse(sort == 'Tex-KLR'& dpi == "d100", 'Tem_1', sort))


# check correlation of clone size based on sorted counts and total counts
clone.df.pheno %>% ggplot(aes(x = CloneSize_Sort, y = CloneSize)) + geom_point() + 
  scale_y_log10() + scale_x_log10() + stat_cor()



```
# select clones that are shared between between timepoints
```{r}
# select clones shared between TP - isolate clones with >5 sorted cells at both timepoints
d21.clones <- clone.df.pheno %>% filter(CloneSize_Sort > 5, dpi == "d21", group == "Ctrl")
d100.clones <- clone.df.pheno %>% filter(CloneSize_Sort > 5, dpi == "d100", group == "Ctrl")
clones.of.interest <- intersect(d21.clones$TRB_CDR3, d100.clones$TRB_CDR3) 

# check number of clones of clones
make_comb_mat(list("d21" = d21.clones$TRB_CDR3, "d100+" = d100.clones$TRB_CDR3))%>% UpSet()

d21.clones$TRB_CDR3 %>% unique() %>% length()

# filter clones based on only those shared
ctrl.df <- clone.df.pheno %>% filter(group == "Ctrl", dpi %in% c("d21", "d100"), TRB_CDR3 %in% clones.of.interest)

# clone size cor over time
ctrl.df %>% group_by(dpi, TRB_CDR3) %>% summarise(Count = sum(Count)) %>% pivot_wider(names_from = dpi, values_from = Count, values_fill = 0) %>% 
  ggplot(aes(x = d21, y = d100)) + geom_point() + geom_smooth(method = "lm", color = "red") + stat_cor()
```

# define clone behaviors at day 100 by kmeans clustering
```{r}
## subset clones of interest at 100 dpi
d100.df <- clone.df.pheno %>% filter(dpi  == "d100", TRB_CDR3 %in% clones.of.interest)

# clonal clustering of clones
d100.wide <- d100.df %>% mutate(freq = Count/CloneSize) %>% select(sort, TRB_CDR3, freq, CloneSize, group) %>% 
  pivot_wider(values_from = freq, names_from = sort, id_cols = c(CloneSize, TRB_CDR3, group)) %>%
  mutate(across(everything(), ~replace_na(.x, 0))) 

set.seed(12)
clusters <- d100.wide %>% select(-c(TRB_CDR3, CloneSize, group)) %>% kmeans(centers = 7)
d100.wide$cluster <- paste0("C", clusters$cluster)

# add cluster based solely on kmeans
d100.wide$cluster[clusters$cluster == "1"] <- "Tex-Prog Bias"
d100.wide$cluster[clusters$cluster == "2"] <- "Tem_2 Bias"
d100.wide$cluster[clusters$cluster == "3"] <- "Tem_1 Bias"
d100.wide$cluster[clusters$cluster == "4"] <- "Divergent"
d100.wide$cluster[clusters$cluster == "5"] <- "Tex-Term Bias"
d100.wide$cluster[clusters$cluster == "6"] <- "Tem_1 Bias"
d100.wide$cluster[clusters$cluster == "7"] <- "Tcm Bias"

# heatmap of clones
right.annot <- rowAnnotation(size = anno_barplot(d100.wide$CloneSize %>% log2))

top.annot <- columnAnnotation(" " = select(d100.wide, -c(TRB_CDR3, CloneSize, cluster, group)) %>% colnames(), 
                              col = list(" " = pop.pal), 
                              show_legend = F)

d100.wide %>% ungroup %>% select(-c(TRB_CDR3, CloneSize, cluster, group)) %>% 
  Heatmap(name = "Frac of Clone", col = brewer.pal(9, "Purples"), 
        split = d100.wide$cluster,
        border = T, row_title_rot = 0, cluster_rows = F,
        width = unit(5*0.6, "cm"), height = unit(6, "cm"), column_dend_height = unit(0.2, "cm"),
        right_annotation = right.annot, top_annotation = top.annot
        )
```

# define clone behaviors at day 21 by kmeans clustering
```{r}
## subset clones of interest at 21 dpi
d021.df <- clone.df.pheno %>% filter(dpi  == "d21", TRB_CDR3 %in% clones.of.interest)

# clonal clustering of clones
d021.wide <- d021.df %>% mutate(freq = Count/CloneSize) %>% select(sort, TRB_CDR3, freq, CloneSize, group) %>% 
  pivot_wider(values_from = freq, names_from = sort, id_cols = c(CloneSize, TRB_CDR3, group)) %>%
  mutate(across(everything(), ~replace_na(.x, 0))) %>% 
  mutate(group = factor(group, levels = c("Ctrl", "aPDL1"))) 



set.seed(12)
clusters <- d021.wide %>% select(-c(TRB_CDR3, CloneSize, group)) %>% kmeans(centers = 4)

d021.wide$cluster[clusters$cluster == "4"] <- "Tex-Int Bias"
d021.wide$cluster[clusters$cluster == "3"] <- "Divergent"
d021.wide$cluster[clusters$cluster == "2"] <- "Tex-Prog Bias"
d021.wide$cluster[clusters$cluster == "1"] <- "Tex-KLR Bias"

# heatmap of clones
right.annot <- rowAnnotation(size = anno_barplot(d021.wide$CloneSize %>% log2))

top.annot <- columnAnnotation(" " = select(d021.wide, `Tex-Int`, `Tex-KLR`, `Tex-Prog`) %>% colnames(), 
                              col = list(" " = pop.pal), 
                              show_legend = F)

d021.wide %>% ungroup %>% select(-c(TRB_CDR3, CloneSize, cluster, group)) %>% 
  Heatmap(name = "Frac of Clone", col = brewer.pal(9, "Purples"), 
        split = d021.wide$cluster,
        border = T, row_title_rot = 0, cluster_rows = T,
        width = unit(3*0.6, "cm"), height = unit(6, "cm"), column_dend_height = unit(0.2, "cm"),
        right_annotation = right.annot,
        top_annotation = top.annot
        )
```

# merge together data for shared clones at d21 and d100
```{r}
# merge day21 and day100 data
longitudinal.df <- merge(
  d021.wide %>% select('TRB_CDR3', 'Tex-Int', 'Tex-KLR', 'Tex-Prog', 'cluster') %>%
    dplyr::rename('d021_Tex-Int' = 'Tex-Int', 'd021_Tex-KLR' = 'Tex-KLR', 'd021_Tex-Prog' = 'Tex-Prog', '21dpi' = 'cluster'),
  d100.wide %>% select('TRB_CDR3', 'Tcm', 'Tem_2', 'Tex-Prog', "Tex-Term", "Tem_1", 'cluster') %>%
    dplyr::rename('d100_Tcm' = 'Tcm', 'd100_Tem_2' = 'Tem_2', 'd100_Tex-Prog' = 'Tex-Prog', 'd100_Tem_1' = 'Tem_1', 'd100_Tex-Term' = 'Tex-Term', '100dpi' = 'cluster')
  ) %>% 
  mutate(`100dpi` = factor(`100dpi`, levels = c("Tcm Bias", "Tex-Prog Bias", "Tex-Term Bias", "Divergent", "Tem_1 Bias", "Tem_2 Bias")), 
         `21dpi` = factor(`21dpi`, levels = c("Tex-Prog Bias", "Divergent", "Tex-Int Bias", "Tex-KLR Bias"))) %>%
  ungroup() %>% arrange(`21dpi`) %>% arrange(`100dpi`)

head(longitudinal.df)

```

# correlation of clone phenotype frequencies at differnt timepoints
```{r}
# calculate correlations of prototypes based on individual clones' intraclonbal freq
cor.input <- longitudinal.df %>% select(-c(TRB_CDR3, `21dpi` , `100dpi`))
head(cor.input)
cor.results <- cor.input %>% as.matrix() %>% rcorr()
cor.results

#check TexProg/cTcm and KLR/Tem1
cor.input %>% ggplot(aes(x = `d021_Tex-Prog` , y =  `d100_Tcm`)) + geom_jitter(size = 0.5, alpha = 0.4) + geom_smooth(method = "lm", color = "red") + stat_cor()
cor.input %>% ggplot(aes(x = `d021_Tex-KLR` , y =  `d100_Tem_1`)) + geom_jitter(size = 0.5, alpha = 0.4) + geom_smooth(method = "lm", color = "red") + stat_cor()
cor.input %>% ggplot(aes(x = `d021_Tex-Prog` , y =  `d100_Tem_1`)) + geom_jitter(size = 0.5, alpha = 0.4) + geom_smooth(method = "lm", color = "red") + stat_cor()

# determine significance
## create new matrix in which to summarize p values
sig_matrix <- matrix(" " , nrow = nrow(cor.results$r), ncol = ncol(cor.results$r))
colnames(sig_matrix) <- colnames(cor.results$r)
rownames(sig_matrix) <- rownames(cor.results$r)

# sumamrize bonferoni corrected pvalues
sig_matrix[cor.results$P <=  0.05 / cor.results$r %>% length()] <- "*"
sig_matrix[cor.results$P <=  0.01 / cor.results$r %>% length()] <- "**" 
sig_matrix[cor.results$P <=  0.001 / cor.results$r %>% length()] <- "***"

# plot all correlations
cor.results$r %>% Heatmap(col =  brewer.pal(9, "Reds"), #viridis(9) , 
                      na_col = "white", border = T, 
                      width = unit(8*0.7, "cm"), height = unit(8*0.7, "cm"), name = "Pearson R", 
                      cluster_rows = T, cluster_columns = T,
                      show_row_dend = T, show_column_dend = T, column_dend_height = unit(0.2, "cm"), row_dend_width = unit(0.2, "cm"), row_dend_side = "right",
                      # bottom_annotation = column.annot, left_annotation = row.annot,
                      # row_title = "Day 21 Behavior" , column_title = "Day 100 Behavior" , 
                      row_title_side = "left", column_title_side = "bottom", row_names_side = "left",
                      cell_fun = function(j, i, x, y, width, height, fill) {grid.text(sig_matrix[i, j], x, y, gp = gpar(fontsize = 14, col = "white"))},
                  )

# plot only different timepoint correlations
column.annot <- columnAnnotation(" " = colnames(cor.results$r[1:3, 4:8]), col = list(" " = pop.pal), show_legend = F)
row.annot <- rowAnnotation(" " = rownames(cor.results$r[1:3, 4:8]), col = list(" " = pop.pal), show_legend = F)

pop.pal

hm.cor <- cor.results$r[1:3, 4:8] %>% Heatmap(
                      
                      col = viridis(9) , 
                      na_col = "white", border = T, 
                      width = unit(5*1, "cm"), height = unit(3*1, "cm"), name = "Pearson R", 
                      cluster_rows = T, cluster_columns = T,
                      show_row_dend = F, show_column_dend = F, 
                      top_annotation = column.annot, right_annotation = row.annot,
                      row_title = "21 dpi Subset" , column_title = "100 dpi Subset" ,
                      row_title_side = "left", column_title_side = "bottom", row_names_side = "left",
                      cell_fun = function(j, i, x, y, width, height, fill) {grid.text(sig_matrix[1:3, 4:8][i, j], x, y, gp = gpar(fontsize = 14, col = "white"))},
                  )

hm.cor
```

# create plots to summarize phenotype transitions
```{r}
## SANKEY PLOT

longitudinal.df %>% make_long(`21dpi`, `100dpi`) %>% 
        ggplot(aes(x = x, next_x = next_x, node = node, next_node = next_node, fill = factor(node), label = node)) +
        geom_sankey(flow.alpha = 0.5, width = .3) + 
        # ggtitle(paste(unique(x$group), unique(x$Longitudinal))) +
        plot.theme + labs(x = element_blank(), y = element_blank()) + 
        theme(axis.line = element_blank(), axis.ticks = element_blank(), axis.text.y = element_blank()) + 
        scale_fill_manual(values = cluster.pal, name = "Clone Behavior")


### HEATMAP

hm1 <- longitudinal.df %>% select(colnames(longitudinal.df)[grepl("d021", colnames(longitudinal.df))]) %>%
      rename_with(~ gsub("d021_", "", .)) %>% 
  Heatmap(name = "Frac of Clone", col = brewer.pal(9, "Purples"), 
      split = longitudinal.df$`21dpi`,
      row_title = paste0(longitudinal.df$TRB_CDR3 %>% length, " Clones Detected 21 and 100+ dpi"),
      cluster_rows = F, cluster_columns = T, show_column_dend = F,
      left_annotation = rowAnnotation("Behavior 21dpi" = longitudinal.df$`21dpi`, col = list("Behavior 21dpi" = cluster.pal), show_legend = F), 
      width = unit(3*0.6, "cm"), height = unit(7, "cm"), 
      column_title = "21dpi",
      border = T)
  
hm2 <- longitudinal.df %>% select(colnames(longitudinal.df)[grepl("d100", colnames(longitudinal.df))]) %>%
      rename_with(~ gsub("d100_", "", .)) %>% 
  Heatmap(name = "Frac of Clone", col = brewer.pal(9, "Purples"), 
      cluster_rows = F, cluster_columns = T, show_column_dend = F,
      left_annotation = rowAnnotation("Behavior 100dpi" = longitudinal.df$`100dpi`, col = list("Behavior 100dpi" = cluster.pal), show_legend = F), 
      width = unit(5*0.6, "cm"), height = unit(7, "cm"), 
      column_title = "100dpi",
      border = T)

hm1 + hm2


```

# logistic regression to determine which d100 behaviors to form cTcm Biased clones
```{r}
# set up dataframe from prediction
regress.df <- longitudinal.df %>% select(TRB_CDR3, `d021_Tex-Int`, `d021_Tex-KLR`, `d021_Tex-Prog`, `21dpi`, `100dpi` ) %>%# take clone bevhaviors and Frac in a given sort d21
  dplyr::rename("TexKLR" = "d021_Tex-KLR", "TexInt" = "d021_Tex-Int", "TexProg" = "d021_Tex-Prog", "Behavior" = "21dpi") %>%
  left_join(d021.wide %>% select(TRB_CDR3, CloneSize)) %>% # add clone size 
  mutate(Tcm = ifelse(`100dpi`== "Tcm Bias", 1, 0)) # if Tcm biased clones -> 1, if not, 0

# Run logistic regression model with Clone_Behavior as the predictor, 
model <- glm(Tcm ~ Behavior, data = regress.df, family = binomial)
summary(model)

# Create new data frame with unique values of Day 21 Behavior and Generate predictions based on Behavior
new_data <- data.frame(Behavior = unique(regress.df$Behavior)) 
predictions <- predict(model, new_data, type = "response", se.fit = TRUE)

# Store predictions and standard errors in the new data frame
new_data$fit <- predictions$fit
new_data$se.fit <- predictions$se.fit

# pairwise comparisions
emm <- emmeans(model, ~ Behavior, )

sig_df <- as.data.frame(pairs(emm)) %>%
  separate(contrast, into = c("group1", "group2"), sep = " - ") %>%
  select(group1, group2, p.value) 

sig_df
  

# plot comparisons
ggplot(data = new_data, aes(x = fct_reorder(Behavior, fit), y = fit, color = Behavior)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey80") +
  geom_point(size = 2.5) + 
  geom_errorbar(aes(ymin = fit - 1.96 * se.fit, ymax = fit + 1.96 * se.fit, width = 0.2), width = 0.25) + # 95% confidence interval
  plot.theme +
  ylab("Predicted Probability of Tcm Bias") +
  xlab("Behavior 21 dpi") +
  theme(legend.position = "none") + coord_flip() + 
  scale_y_continuous(limits = c(0,1.01)) +
  scale_color_manual(values = cluster.pal)


```







