# Pre-processing for scRNA/TCR-seq of GP33+ CD8 T cells 100+ dpi of LCMV-Cl13 with and without aPDL1 Treatment ftom 22-25dpi
## pre-processing starts with outs from cellranger - available on Github and GEO (GSE285412)
## file names and paths match the original IDs use in analysis and will need to be updated to match uploaded IDs

# load common libraries
```{r}
suppressPackageStartupMessages({
    library(tidyverse) 
    library(Seurat)
    library(RColorBrewer)
    library(viridis)
    library(cowplot)
    library(ggridges)
    library(data.table)
    library(ggpubr)
  })

path <- "~/Experiments/CR120_scRNATCR_aPDL1"
outs <- paste0(path, "/01_FilterMultiseqIntegration/outs")

```

# helper functions for generating seurat object
```{r}
###TCR Helper Functions for Creating Seurat Object

###TCR Helper Functions

# ==== color palettes ====
# TCR chains captured palette
tcr_chains_pal <- c("#006080","#008EA0FF", 
                    "#6BAED6", "#C6DBEF", "#238B45", "#74C476",
                    "#FF6F00FF", "#C71000FF", "#b8b8b899")
names(tcr_chains_pal) <- c("ab", "aab", 
                           "a", "aa", "b", "bb", 
                           "abb", "aabb", "none")


# ==== HELPER FUNCTIONS: SEURAT OBJECT CREATION ====

extract_tcr_chains <- function(tcr_data, aa=TRUE, nt=TRUE) {
  
  if (aa) {
    # check that required cdr3 columns are present
    if (! "cdr3s_aa" %in% colnames(tcr_data)) {
      stop("cdr3s_aa column not found in data")
    }
    
    # split cdr3s into individual chains (aa)
    tcr_data$TRA_aa <- gsub(".*?TRA:(\\w+);*.*", "\\1", tcr_data$cdr3s_aa)
    tcr_data$TRA_aa[grepl("TRB:", tcr_data$TRA_aa)] <- NA
    
    tcr_data$TRA_secondary_aa <- gsub(".*?TRA:.*;TRA:(\\w+).*", "\\1", tcr_data$cdr3s_aa)
    tcr_data$TRA_secondary_aa[grepl("TR[AB]:", tcr_data$TRA_secondary_aa)] <- NA
    
    tcr_data$TRB_aa <- gsub(".*?TRB:(\\w+);*.*", "\\1", tcr_data$cdr3s_aa)
    tcr_data$TRB_aa[grepl("TRA:", tcr_data$TRB_aa)] <- NA
    
    tcr_data$TRB_secondary_aa <- gsub(".*TRB:.*TRB:(\\w+);*.*", "\\1", tcr_data$cdr3s_aa)
    tcr_data$TRB_secondary_aa[grepl("TR[AB]:", tcr_data$TRB_secondary_aa)] <- NA
  }
  
  if (nt) {
    # check that required cdr3 columns are present
    if (! "cdr3s_nt" %in% colnames(tcr_data)) {
      stop("cdr3s_aa column not found in data")
    }
    
    # split cdr3s into individual chains (nt)
    tcr_data$TRA_nt <- gsub(".*?TRA:(\\w+);*.*", "\\1", tcr_data$cdr3s_nt)
    tcr_data$TRA_nt[grepl("TRB:", tcr_data$TRA_nt)] <- NA
    
    tcr_data$TRA_secondary_nt <- gsub(".*?TRA:.*;TRA:(\\w+).*", "\\1", tcr_data$cdr3s_nt)
    tcr_data$TRA_secondary_nt[grepl("TR[AB]:", tcr_data$TRA_secondary_nt)] <- NA
    
    tcr_data$TRB_nt <- gsub(".*?TRB:(\\w+);*.*", "\\1", tcr_data$cdr3s_nt)
    tcr_data$TRB_nt[grepl("TRA:", tcr_data$TRB_nt)] <- NA
    
    tcr_data$TRB_secondary_nt <- gsub(".*TRB:.*TRB:(\\w+);*.*", "\\1", tcr_data$cdr3s_nt)
    tcr_data$TRB_secondary_nt[grepl("TR[AB]:", tcr_data$TRB_secondary_nt)] <- NA
  }
  
  return(tcr_data)
}

# extract chain combination (i.e. alpha, beta, alpha/beta, alpha/alpha/beta, etc.)
count_chains <- function(cdr3s_aa) {
  
  if (is.na(cdr3s_aa)) {
    return(NA)
  }
  
  a_count <- stringr::str_count(cdr3s_aa, "TRA:")
  b_count <- stringr::str_count(cdr3s_aa, "TRB:")
  
  chains <- paste(c(rep("a", a_count), rep("b", b_count)), collapse="")
  
  return(chains)
}

# merge cellranger vdj clone annotations with clonotypes
add_gene_annot <- function(chain, clonotypes, clonotype_annot) {
  xcols <- c("clonotype_id", paste0(chain, "_aa"))
  ycols <- c("clonotype_id", "cdr3")
  
  clonotypes <- dplyr::left_join(clonotypes, clonotype_annot[,c("clonotype_id", "v_gene", "d_gene", 
                                                                "j_gene", "c_gene", "cdr3")],
                                 by = setNames(ycols, xcols)
  )
  
  ncols <- ncol(clonotypes)
  colnames(clonotypes)[(ncols-3):ncols] <- paste(chain, colnames(clonotypes)[(ncols-3):ncols], sep="_")
  
  return(clonotypes)
}

# calculate proportion of cells with TCR chains captured
tcr_capture_rate <- function(obj, group_var, facet_var=NULL) {
  
  # convert to metadata in case input is seurat data object
  # enables function to accept either seurat obj, seurat metadata, or other dataframe
  if(class(obj) == "Seurat") {
    obj <- obj[[]]
  }
  
  if (is.null(obj$cdr3s_aa)) {
    print("no cdr3s_aa column")
    return(NULL)
  }
  
  # check TCR chains captured exists in dataframe
  # if not, calculate
  if (! "chains" %in% colnames(obj)) {
    obj$chains <- sapply(obj$cdr3s_aa, count_chains)
  }
  
  if (!is.null(facet_var)) {
    groups <- c(group_var, facet_var, "chains")
  } else {
    groups <- c(group_var, "chains")
  }
  tcr_capture_df <- obj %>% group_by_at(groups) %>% 
    dplyr::summarize(num_cells=n())
  
  tcr_capture_df$chains[is.na(tcr_capture_df$chains)] <- "none"
  tcr_capture_df$chains <- factor(tcr_capture_df$chains, 
                                  levels=names(tcr_chains_pal))
  
  p <- ggplot(tcr_capture_df, aes_string(x=group_var, y="num_cells", fill="chains")) +
    geom_bar(stat="identity", position = position_fill(reverse = TRUE)) + 
    coord_cartesian(expand=F) +
    scale_fill_manual(values=tcr_chains_pal, na.value="grey", name="TCR chains") +
    labs(y="Proportion of cells") + 
    theme_classic() 
  
  if (!is.null(facet_var)) {
    p <- p + facet_grid(as.formula(paste("~", facet_var)), space="free_x", scales="free_x")
  }
  
  res <- list(plot=p, data=tcr_capture_df)
  return(res)
}
```

# generate seurat objects from cell ranger outs

```{r}

lanes <- c("CR120_001" , "CR120_002", "CR120_003")

# run function that creates a so for each 10X lane

so.list <- lapply(lanes, function(x){

  # define lane_path for the inidividual lane
  lanepath_path <- paste0(path, "/00_CellRangerOuts/" , x)
  
  
  # generate counts matrix
  counts <- Read10X_h5(paste0(lanepath_path , "/filtered_feature_bc_matrix.h5"))
  
  # Append TCR information to metadata tables
  tcr_table <- fread(input = paste0(lanepath_path, "/filtered_contig_annotations.csv"))
  
  clonotype_table <- fread(input = paste0(lanepath_path, "/clonotypes.csv"))
  
  # definte extract_tcr_chains
  clonotype_table <- extract_tcr_chains(clonotype_table, aa=TRUE, nt=TRUE)
  
  #unique tcrs
  consensus_table <- unique(tcr_table[,c("raw_clonotype_id", "chain", "cdr3", "cdr3_nt", "v_gene", "d_gene", "j_gene", "c_gene")])
  colnames(consensus_table)[1] <- "clonotype_id" 
  
  # create dataframe with all clonotype information 
  chains <- c("TRA", "TRB", "TRA_secondary", "TRB_secondary")
  
  for (chain in chains) {
    clonotype_table <- add_gene_annot(chain, clonotype_table, consensus_table)
  }
  
  clonotype_table <- clonotype_table[!duplicated(clonotype_table$clonotype_id)]
  
  tcr <- tcr_table[,c("barcode", "raw_clonotype_id")]
  names(tcr)[names(tcr) == "raw_clonotype_id"] <- "clonotype_id"
  tcr <- tcr[!duplicated(tcr$barcode), ]
  #add tcr data to dataframe with all
  tcr_data <- merge(tcr, clonotype_table, by="clonotype_id", all.x = TRUE)
  # TCR chains captured
  tcr_data$chains <- sapply(tcr_data$cdr3s_aa, count_chains)
  
  rownames(tcr_data) <- tcr_data$barcode
  
  so <- CreateSeuratObject(counts = counts, meta.data = tcr_data)
  
  so$orig.ident <- x
  
  return(so)

})

so.list

```

# tcr QC

```{r}

tcr.plots <- lapply(so.list, function(x){
  tcr_capture_res <- tcr_capture_rate(x, group_var = "orig.ident")
  return(tcr_capture_res$plot + xlab(element_blank()))
})

do.call(ggarrange , c(tcr.plots, common.legend = T, legend = "right" , ncol = 3))

```

# add percent MT  to seurat object
```{r}
# add percent mt 
so.list <- lapply(so.list, function(so){
  so$percent.mt <- PercentageFeatureSet(so, pattern = "^mt-")
  return(so)
})
```

# initial filtering of SO based on MT and features
```{r}
# calculate cutoffs for % mito and n feature - based on 2x SD for features and 95%il for %mito
nfeature_cutoffs_list <- lapply(so.list, 
                                function(x) {
                                  data.frame(orig.ident=unique(x$orig.ident),
                                             nfeature_min=max(1000, mean(x$nFeature_RNA) - 2*sd(x$nFeature_RNA)),
                                             nfeature_max=mean(x$nFeature_RNA) + 2*sd(x$nFeature_RNA),
                                             mito=min(4, quantile(x$percent.mt, 0.95))
                                            )
                                })
nfeature_cutoffs <- do.call(rbind, nfeature_cutoffs_list)
rownames(nfeature_cutoffs) <- nfeature_cutoffs$orig.ident

nfeature_cutoffs

write.csv(nfeature_cutoffs, file = paste0(outs, "/nfeaturecutoffs.csv"))

# library QC metrics
qc_metrics_list <- lapply(so.list, 
                          function(x) x[[c("orig.ident", "nFeature_RNA", "nCount_RNA", "percent.mt", "Phase" , "chains")]])

qc_metrics <- do.call(rbind, qc_metrics_list)

p1 <- ggplot(qc_metrics, aes(x=percent.mt, y = nFeature_RNA))  + 
  geom_point(size = 0.1) + 
  stat_density_2d(aes(fill = ..level..), geom = "polygon", show.legend = F) +
  facet_wrap(~orig.ident) + scale_x_log10() +
  scale_fill_viridis(option = "magma") +
  geom_rect(data = nfeature_cutoffs ,
           aes(x = mito, xmin = 0, xmax = mito,
               y = nfeature_min, ymin = nfeature_min,  ymax= nfeature_max),
           colour="red", lty=2, size=0.5, fill = "transparent")

p3 <- ggplot(qc_metrics, aes(x=percent.mt, y = nFeature_RNA, color = chains))  + 
  geom_point(size = 0.1) + 
  #stat_density_2d(aes(fill = ..level..), geom = "polygon", show.legend = F) +
  facet_wrap(~orig.ident) + scale_x_log10() +
  scale_color_manual(values=tcr_chains_pal, na.value="grey", name="TCR chains")

plot_grid(p1, p3, ncol = 2)

ggsave(paste(outs, "nFeature_Cutoffs.pdf" , sep = "/"), width = 12)

# set the tcr combinatios which shouldnt exist
bad_tcr_combs <- c("aaaab", "aaabb","aaaaab","aaaabb","aaaabbb","aaab", "aaabbb","aabb", "aabbb", "abb", "abbb","aabbbb","abbbb")

# filter seruat objects
filtered.so.list <- lapply(so.list, function(so) {
  # define feature cutoffs for particular 10X lane
  cut_offs <- nfeature_cutoffs %>% subset(orig.ident == so$orig.ident %>% unique) 
  # define cells to keep 
  so$keep <- so$nFeature_RNA > cut_offs$nfeature_min & 
    so$nFeature_RNA < cut_offs$nfeature_max & 
    so$percent.mt < cut_offs$mito & 
    (!so$chains %in% bad_tcr_combs) 
  so.filt <- subset(so, keep==TRUE)
  return(so.filt)
})

# of cells in each lane
lapply(filtered.so.list, function(so) {table(so$orig.ident)}) 
 

```

# demultiplexing - function
```{r}
ms.demux <- function(so) {
  require(deMULTIplex2)
  require(deMULTIplex)

# create read table ++++++++++
  # get the sample name out of orig.ident 
  # lane.id <- so$orig.ident %>% substr(0,3) %>% unique()
  # lane.name <- so$orig.ident %>% substr(5,100) %>% unique()
  # name.id.name <- paste(lane.name, lane.id, sep = "_")
  lane.id <- so$orig.ident %>% unique()

  # get fastq file name base
  files <- list.files(ms.fastq) 
  ms.file.root <- files[grepl(lane.id, files)][1] %>% # take first fastq file
      str_replace("(L\\d{3}).*", "\\1") # remove all text after the L00X number

  read_table <- readTags(dir = ms.fastq,
                       name = ms.file.root, ### CHANGE THIS
                       barcode.type = "MULTIseq",
                       assay = "RNA")


# create read matrix
  tag_mtx <- alignTags(read_table, tag.ref) 
  
  cell_ids <- rownames(tag_mtx) %in% substr(rownames(so@meta.data), 1, 16) # choose only cells in so
  tag_used <- Matrix::colSums(tag_mtx) > 1e4 # Filter for used tags
  
  tag_mtx_filtered <- tag_mtx[cell_ids,tag_used]

# barcode classification ++++++++++
  bar.table <- tag_mtx_filtered %>% data.frame %>% dplyr::select(c(1:nbar))
  counter <- 0
  while (counter >= 0) {
    counter <- counter + 1
    print(paste("Round ", counter, "...",sep=""))
    bar.table_sweep.list <- list()
    n <- 0
    for (q in seq(0.01, 0.99, by=0.02)) {
      print(q)
      n <- n + 1
      bar.table_sweep.list[[n]] <- classifyCells(bar.table, q=q)
      names(bar.table_sweep.list)[n] <- paste("q=",q,sep="")
    }
    res <- findThresh(call.list=bar.table_sweep.list)
    calls <- classifyCells(bar.table, q=findQ(res$res, res$extrema))
    negs <- names(calls)[which(calls == "Negative")]
    if (length(negs) == 0) { break }
    if (counter == 1) { neg.cells <- negs }
    if (counter > 1) { neg.cells <- c(neg.cells, negs) }
    print(paste(length(which(calls %in% c("Negative","Doublet"))), " Singlets...", sep=""))
    print(paste(length(which(calls == "Doublet")), " Negatives...", sep=""))
    print(paste(length(which(calls == "Negative")), " Negatives...", sep=""))
    bar.table <- bar.table[-which(rownames(bar.table) %in% neg.cells), ]
  }

  call.table <- data.frame(names(calls), calls)
  colnames(call.table) <- c("Cell_Barcode", "Call")
  
  
  
# vsiualization ++++++++++
  
  # dim reduction 
  bar.tsne <- barTSNE(bar.table[,1:nbar]) #these values need to be changed based on how many barcodes you used
  bar.tsne$Cell_Barcode <- rownames(bar.tsne)
  
  # merge together clustering and dim reduction
  ms.data <- full_join(bar.tsne, call.table, by = "Cell_Barcode")

  
  # plots
  tsne <- ms.data %>% ggplot(aes(x = TSNE1, y = TSNE2, color = Call)) + geom_point(size = 0.3)
  
  features <- ms.data %>% dplyr::select(-c("Cell_Barcode", "Call")) %>%
    pivot_longer(cols = !c(TSNE1, TSNE2)) %>%
    ggplot(aes(x = TSNE1, y = TSNE2, color = value)) + geom_point(size = 0.3) + facet_wrap(~name) + 
    scale_color_viridis_c()
  
  plot_grid(tsne , features, rel_widths = c(2.5,3))

# save export ++++++++++
  ggsave(path = paste0(outs,"/multiseq"), filename = paste("MultiSeqTSNE_" , unique(so$orig.ident), ".pdf") , height = 5, width = 12)
  write.csv(ms.data %>% dplyr::select(c(Cell_Barcode, Call)),
            file = paste0(outs, "/multiseq//MultiSeqCalls_" , unique(so$orig.ident), ".csv"))
  write.csv(tag_mtx_filtered, file = paste0(outs, "/multiseq/MultiSeqBarTable_" , unique(so$orig.ident), ".csv"))

# add to new SO ++++++++++
  so <- AddMetaData(object = so, metadata = setNames(ms.data$Call, paste0(ms.data$Cell_Barcode, "-1")), col.name =  "multiseq_class") 

return(so)

}

```

# demux all lanes with above function
```{r}

# set up for multiseq demuxing
bar.ref.full <- read.csv(paste0(path, "/00_CellRangerOuts/C120_Bartable.csv"))  #refernce for barcodes
bar.ref.full

tag.ref <- c(bar.ref.full$Barcode.Seuqence) # pull out just bar codes
names(tag.ref) <- bar.ref.full$Sample # give names to make demuxing way easier
# tag.ref <- tag.ref[1:11] # choose only bars that we actually used - change this
nbar <- length(tag.ref)
ms.fastq <- paste0(path, "/00_MS_FASTQ/") # path for fastq files
ms.fastq

# run function to demultiplex
so.list.demuxed <- filtered.so.list %>% lapply(ms.demux)
```


# create list of seurat objects
```{r}
so.list.demuxed

qc_metrix <- lapply(so.list.demuxed, function(x) x[[c("orig.ident", "nFeature_RNA", "nCount_RNA", "percent.mt", "Phase" , "chains", "multiseq_class")]])

qc_metrix <- do.call(rbind, qc_metrix)

# tcr chains

qc_metrix %>% ggplot(aes(x = multiseq_class, fill = chains)) + 
  geom_bar() + 
  facet_wrap(~orig.ident) +
  rotate_x_text(90) + scale_fill_manual(values = tcr_chains_pal) + 
  xlab("Multiseq Classification")

# % mito and nfeature

p1 <- qc_metrix %>% ggplot(aes(y = nFeature_RNA, x = multiseq_class)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(size = 0.3, width = 0.2, alpha = 0.5) + 
  rotate_x_text(45) + xlab("Multiseq Classification")

p2 <-qc_metrix %>% ggplot(aes(y = percent.mt, x = multiseq_class)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(size = 0.3, width = 0.2, alpha = 0.5) + 
  rotate_x_text(45) + xlab("Multiseq Classification")

ggarrange(p1, p2, common.legend = T, legend  = "right")

ggplot(qc_metrix, aes(x=percent.mt, y = nFeature_RNA))  + 
  geom_point(size = 0.1, aes(color = multiseq_class)) + 
  facet_wrap(~orig.ident) 

table(qc_metrix$orig.ident, qc_metrix$multiseq_class)
```

# save RDS
```{r}
saveRDS(so.list.demuxed, file = "~/Experiments/CR120_scRNATCR_aPDL1/99_seuratobjects/8A_aPDL1_demuxed_so_list.rds")
```
