# In the markdown, we integrate the Seurat object from this experiment along with Data From the scRNA/TCR atlas in ***Figure 1***. 
### We do this bevause cTem are such a large portion of the Gp33+ at a late timepoint that clustering analysis become difficlult to interpret as you wind up with many cTem clusters before you start to differentiate other populations from one another

# load common libraries + set paths + import seurat datasets
```{r}
suppressPackageStartupMessages({
  library(tidyverse) 
  library(Seurat)
  library(RColorBrewer)
  library(viridis)
  library(cowplot)
  library(ggridges)
  library(data.table)
  library(ggpubr)
  library(ComplexHeatmap)
})

path <- "~/Experiments/CR120_scRNATCR_aPDL1"
outs <- paste0(path, "/01_FilterMultiseqIntegration/outs")

so.list.cr119 <- read_rds("~/ChronicMemoryGithub_Submission/1_scRNATCRseqAtlas/data/1A_DemuxedSeuratObjectList_batch2.rds") # split by CD62L
so.list.cr082 <- read_rds("~/ChronicMemoryGithub_Submission/1_scRNATCRseqAtlas/data/1B_DemuxedSeuratObjectList_batch1.rds") # not split by CD62L
so.list.cr120 <- read_rds("~/ChronicMemoryGithub_Submission/8_aPDL1scRNATCR/data/8A_aPDL1_demuxed_so_list.rds") # this experiment
```

# remove genes function
```{r}
#remove TCR genes from VF lists
remove_genes <- function(features, species, tcr=TRUE, ig=TRUE, cell_cycle=TRUE, mito=TRUE, histone=TRUE, ribosome=TRUE) {
    print(paste("Before filtering:", length(features), "features"))

    if (species == "human") {
        if (tcr) {
            features <- features[!grepl("^TR[ABGD][VDJC]", features)]
        }
        if (ig) {
            features <- features[!grepl("^IG[HLK][VDJC]", features)]
        }
        if (cell_cycle) {
            features <- features[!(features %in% unlist(cc.genes))]
        }
        if (mito) {
            features <- features[!grepl("^MT-", features)]
        }
    }
    if (species == "mouse") {
        if (tcr) {
            features <- features[!grepl("^Tr[abgd][vdjc]", features)]
        }
        if (tcr) {
            features <- features[!grepl("^Tcr[abgd][vdjc]", features)]
        }
        if (ig) {
            features <- features[!grepl("^Ig[hlk][vdjc]", features)]
        }
        if (cell_cycle) {
            features <- features[!(features %in% stringr::str_to_title(unlist(cc.genes)))]
        }
        if (mito) {
            features <- features[!grepl("^Mt-", features)]
        }
      
       if (histone) {
            features <- features[!grepl("^Hist", features)]
       }
      
       if (ribosome) {
            features <- features[!grepl("^Rps", features)]
       }
      
      if (ribosome) {
            features <- features[!grepl("^Rpl", features)]
        }
      
    }

    print(paste("After filtering:", length(features), "features"))
    
    return(features)
}
```

# select GP33+ 10X wells only | rename metadata for seruat objects
```{r}
# CR82  +++++++++++++++++++++++++ +++++++++++++++++++++++++ +++++++++++++++++++++++++ +++++++++++++++++++++++++ +++++++++++++++++++++++++ +++++++++++++++++++++++++
so.list.cr082.gp33 <-  list(so.list.cr082[[1]], so.list.cr082[[2]], so.list.cr082[[3]])

so.list.cr082.gp33 <- so.list.cr082.gp33 %>% lapply(function(so){ 
 so@meta.data <- so@meta.data %>% mutate(group = paste(orig.ident, multiseq_class, sep = "_")) %>%
    separate(group, into = c("Tet", "LaneID", "Mouse")) %>%
    mutate(Tet = substr(Tet, 0,4)) %>%
    mutate(Sort = "Total") %>%
    mutate(LCMV = ifelse(grepl( "Chron", Mouse), "Cl13", NA)) %>% mutate(LCMV = ifelse(grepl("Naive", Mouse), "Naive", LCMV)) %>%
        mutate(LCMV = ifelse(grepl("Acu", Mouse), "Arm", LCMV)) %>%
    mutate(DPI = ifelse(grepl("Chron", Mouse) & !grepl("ChronMem", Mouse), "d26", NA)) %>% mutate(DPI = ifelse(grepl("ChronMem", Mouse), "d100", DPI)) %>%
      mutate(DPI = ifelse(grepl("Acu", Mouse), "d69", DPI)) %>%
    mutate(Rep = ifelse(Mouse != "Naive" & Mouse != "Doublet", substr(Mouse, nchar(Mouse), nchar(Mouse)), NA)) %>%
    mutate(Group = ifelse(LCMV == "Naive", "Naive", paste(LCMV, DPI, Tet, sep = "_"))) %>%
    mutate(Group = ifelse(LCMV == "Naive", "Naive", paste(LCMV, DPI, Tet, Sort, sep = "_"))) %>% select(-Mouse) %>% select(-LaneID) %>% 
    mutate(orig.ident = "CR82")
  return(so)
  })

# so.list.cr082.gp33[[1]] %>% head

# CR119 +++++++++++++++++++++++++ +++++++++++++++++++++++++ +++++++++++++++++++++++++ +++++++++++++++++++++++++ +++++++++++++++++++++++++ +++++++++++++++++++++++++

so.list.cr119.gp33 <- list(so.list.cr119[[1]], so.list.cr119[[3]], so.list.cr119[[4]], so.list.cr119[[5]])

so.list.cr119.gp33 <- so.list.cr119.gp33 %>% lapply(function(so){ 
  so@meta.data <- so@meta.data %>% mutate(group = paste(orig.ident, multiseq_class, sep = "_")) %>% 
    separate(group, into = c("LaneID", "Tet", "Sort", "LCMV", "DPI", "Rep")) %>% 
    select(-LaneID) %>% 
    mutate(Group = ifelse(LCMV == "Naive", "Naive", paste(LCMV, DPI, Tet, sep = "_"))) %>%
    mutate(Group = ifelse(LCMV == "Naive", "Naive", paste(LCMV, DPI, Tet, Sort, sep = "_"))) %>% 
    mutate(orig.ident = "CR119")
  return(so)
  })


# CR120 +++++++++++++++++++++++++ +++++++++++++++++++++++++ +++++++++++++++++++++++++ +++++++++++++++++++++++++ +++++++++++++++++++++++++ +++++++++++++++++++++++++

# so.list.cr120_2 <- list(so.list.cr120[[1]], so.list.cr120[[2]], so.list.cr120[[3]])

# so.list.cr120_2[[1]] %>% class


so.list.cr120 <- so.list.cr120 %>% lapply(function(so){ 
  so@meta.data <- so@meta.data %>% 
    separate(multiseq_class, into = c("Group", "Rep"), remove = F) %>%
    mutate(orig.ident = "CR120", LCMV = "Chron", DPI = "d100", Tet = "GP33", Sort = "Total") 
  return(so)
  })

# so.list.cr120_2[[1]] %>% class

# create a list of all objects
so.list <- c(so.list.cr119.gp33[[1]], so.list.cr119.gp33[[2]], so.list.cr119.gp33[[3]], so.list.cr119.gp33[[4]],
            so.list.cr082.gp33[[1]], so.list.cr082.gp33[[2]], so.list.cr082.gp33[[3]], 
            so.list.cr120[[1]], so.list.cr120[[2]], so.list.cr120[[3]]) 


```

# pre-processing
```{r}
so.list <- lapply(so.list, function(x) {
  print(x$orig.ident %>% unique())
  # remove doublets and unidentified cells from individual so w/in so.list
  x <- subset(x, !is.na(multiseq_class) & multiseq_class != "Doublet")
  x <- NormalizeData(x)
  x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
  return(x)
})
```

# integration workflow: seurat reciprocal PCA
```{r}

## find overlapping features
### remove TCR genes etc
features <- SelectIntegrationFeatures(object.list = so.list) %>% 
  remove_genes(species = "mouse" , tcr=T, ig=F, cell_cycle=F, mito=F, histone=F, ribosome=F)

## PCA run on integration features, not all features
so.list <- lapply(so.list, FUN = function(x) {
    x <- ScaleData(x, features = features, verbose = F)
    x <- RunPCA(x, features = features, verbose = F)
})

## ID anchors
anchors <- FindIntegrationAnchors(object.list = so.list, anchor.features = features, reduction = "rpca")#, k.anchor = 5)

## created integrated dataset
so.rpca <- IntegrateData(anchorset = anchors)

## specify that we will perform downstream analysis on the corrected data note that the
### original unmodified data still resides in the 'RNA' assay
DefaultAssay(so.rpca) <- "integrated"

```

# scale gene expression and add cell cycles
```{r}
so.rpca <- ScaleData(so.rpca)
so.rpca <- RunPCA(so.rpca, npcs = 30, features = VariableFeatures(so.rpca)) # only perfrom PCA on cleaned integration features

# add cell cycle
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes
so.rpca <- CellCycleScoring(so.rpca, s.features = s.genes, g2m.features = g2m.genes)

```

# cluster the integrated object
```{r}
so.clusters.1 <- RunUMAP(so.rpca, dims = 1:12)
DefaultAssay(so.clusters.1) <- "integrated"
so.clusters.1 <- FindNeighbors(so.clusters.1, dims = 1:12)
so.clusters.1 <- FindClusters(so.clusters.1,resolution = 0.5)

so.clusters.1@meta.data <- so.clusters.1@meta.data %>%
  mutate(Group2 = ifelse(LCMV == "Arm", "Arm", NA)) %>%
  mutate(Group2 = ifelse(LCMV %in% c("Cl13", "Chron") & DPI == "d100", "Cl13 d100+", Group2)) %>%
  mutate(Group2 = ifelse(LCMV %in% c("Cl13", "Chron") & DPI != "d100", "Cl13 d21-26", Group2)) %>%
  mutate(Group2 = ifelse(LCMV == "Naive", "Naive", Group2)) %>%
  mutate(Group2 = ifelse(Group == "Ctrl", "PBS", Group2)) %>% mutate(Group2 = ifelse(Group == "aPDL1", "aPDL1", Group2))


DimPlot(so.clusters.1, label = T)
DimPlot(so.clusters.1, split.by = "Group2", ncol = 2) & NoAxes()

DefaultAssay(so.clusters.1) <- "RNA"
FeaturePlot(so.clusters.1, features = c("Klrg1", "Cx3cr1", "Pdcd1", "Tox", "Cd101" , "Tcf7", "Sell", "Cxcr3"), ncol = 4) & NoLegend() & NoAxes()
DimPlot(so.clusters.1, split.by = "Group", ncol = 2, group.by = "orig.ident", shuffle = T) & NoAxes()

DimPlot(so.clusters.1, label = T)
FeaturePlot(so.clusters.1, features = c("percent.mt", "nCount_RNA", "nFeature_RNA")) & NoAxes()
VlnPlot(so.clusters.1, features = c("percent.mt", "nCount_RNA", "nFeature_RNA"))
DimPlot(so.clusters.1, group.by = "Phase") & NoAxes()
DimPlot(so.clusters.1, group.by = "Group", shuffle = T) & NoAxes()


# calculate DEG
cluster.markers <- FindAllMarkers(so.clusters.1, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25, assay = "RNA", features = features)

# get just the top 5 featurs per cluster
cluster.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> topgenes

hm.df <- DotPlot(so.clusters.1, features=unique(topgenes$gene))$data[, c("features.plot", "id", "pct.exp", "avg.exp.scaled")]
## mutate the data bit
hm.df <- hm.df[,c(1,2,4)] %>% pivot_wider(names_from = id, values_from = avg.exp.scaled)%>% as.data.frame
rownames(hm.df) <- hm.df$features.plot

## make heatmap
hm <- hm.df[,-1] %>% 
  Heatmap(name = "Z-score", # name over the legend
          col = rev(brewer.pal(9, "RdBu")), # values for the cell values
          height = unit(16, "in"), width = unit(3, "in"), 
          show_column_dend = F, show_row_names = T, column_dend_height = unit(0.15, "in"), 
          cluster_rows = T, cluster_columns = T # do you want the rows and columns to cluster
        )

# pdf(file = paste0(outs, "/1C_Clusters_1.pdf"), height = unit(18, "in"), width = unit(6, "in"))
hm
# dev.off()

DimPlot(so.clusters.1, label = T)
# ggsave("1C_UMAP_1.pdf", path = outs)

```


# remove junk cells (C12,13) and naive cells (C5) and peform a second round of clustering
```{r}
# set bad clusters
### c5 = naive
### c13 = likely doublets -  express KLRG1 and Sell
### c12 = high mito - high nfeature

so.clusters.1$keep <- "yes"
so.clusters.1$keep[so.clusters.1$seurat_clusters %in% c("5", "12", "13")] <- "no"
so.clusters.1$keep[so.clusters.1$multiseq_class == "Naive"] <- "no"
DimPlot(so.clusters.1, group.by = "keep")

# susbet seurat obj
so.clusters.2 <- subset(so.clusters.1, keep == "yes")

# seurat processing
DefaultAssay(so.clusters.2) <- "integrated"
so.clusters.2 <- FindVariableFeatures(so.clusters.2)

features.2 <- remove_genes(VariableFeatures(so.clusters.2),
                         species = "mouse" , tcr=T, ig=F, cell_cycle=T, mito=T, histone=T, ribosome=T)

so.clusters.2 <- ScaleData(so.clusters.2, verbose = T)
so.clusters.2 <- RunPCA(so.clusters.2, npcs = 30, verbose = T, features = features.2)

# clustering & UMAP
so.clusters.2 <- FindNeighbors(so.clusters.2, dims = 1:15)
so.clusters.2 <- FindClusters(so.clusters.2,resolution = 0.34)
so.clusters.2 <- RunUMAP(so.clusters.2, dims = 1:15, return.model = T)

# find DEG
cluster.markers <- FindAllMarkers(so.clusters.2, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25, assay = "RNA", features = features)

cluster.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> topgenes
hm.df <- DotPlot(so.clusters.2, features=unique(topgenes$gene))$data[, c("features.plot", "id", "pct.exp", "avg.exp.scaled")]
hm.df <- hm.df[,c(1,2,4)] %>% pivot_wider(names_from = id, values_from = avg.exp.scaled)%>% as.data.frame
rownames(hm.df) <- hm.df$features.plot

hm <- hm.df[,-1] %>%
  Heatmap(name = "Z-score", # name over the legend
          col = rev(brewer.pal(9, "RdBu")), # values for the cell values
          height = unit(16, "in"), width = unit(3, "in"),
          show_column_dend = F, show_row_names = T, column_dend_height = unit(0.15, "in"),
          cluster_rows = T, cluster_columns = T # do you want the rows and columns to cluster
        )

pdf(file = paste0(outs, "/1C_Clusters_2.pdf"), height = unit(18, "in"), width = unit(6, "in"))
hm
dev.off()

DimPlot(so.clusters.2, label = T)
# ggsave("1C_UMAP_2.pdf", path = outs)

DimPlot(so.clusters.2, split.by = "Group2", ncol = 3, label = T) & NoAxes()

DefaultAssay(so.clusters.2) <- "RNA"
FeaturePlot(so.clusters.2, features = c("Klrg1", "Cx3cr1", "Pdcd1", "Tox", "Cd101" , "Tcf7", "Sell", "Cxcr3"), ncol = 4) & NoLegend() & NoAxes()
VlnPlot(so.clusters.2, features = c("percent.mt", "nCount_RNA", "nFeature_RNA"), pt.size = 0)
```

# rename clusters and save seruat object
```{r}
so.clusters.2@meta.data <- so.clusters.2@meta.data %>% 
  mutate(cell_cluster = case_when(
    seurat_clusters == "0"  ~ "Tem_1",
    seurat_clusters == "1"  ~ "Tcm",
    seurat_clusters == "2"  ~ "Tex-Term_1",
    seurat_clusters == "3"  ~ "Tex-KLR",
    seurat_clusters == "4"  ~ "Tem_2",
    seurat_clusters == "5"  ~ "Tex-Prog",
    seurat_clusters == "6"  ~ "Tex-Int",
    seurat_clusters == "7"  ~ "Tex-Term_2",
    seurat_clusters == "8"  ~ "Prolif",
    seurat_clusters == "9"  ~ "Tex-ISG"
  ))

Idents(so.clusters.2) <- factor(so.clusters.2$cell_cluster,
                           levels = c( "Tcm", "Tem_1", "Tem_2", "Tex-KLR", "Tex-Int", "Tex-Term_1", "Tex-Term_2", "Tex-Prog",  "Tex-ISG", "Prolif"))

DimPlot(so.clusters.2, label = T)

saveRDS(so.clusters.2, file = paste0(path, "/99_seuratobjects/8B_aPDL1_integratedseuratobj.rds"))


```


