# here, we are perfoming analysis to make all plots associated with this dataset. we will look at deg between clusters, compare gene expression state of cells from both groups, and do analysis of clonal behaviors

# load in seurat object and commonly used packages
```{r}
suppressPackageStartupMessages({
  library(tidyverse) 
  library(Seurat)
  library(RColorBrewer)
  library(viridis)
  library(cowplot)
  library(ggridges)
  library(data.table)
  library(ggpubr)
  library(ComplexHeatmap)
  library(ggrepel)
  library(writexl)
  library(Hmisc)
})

path <- "~/ChronicMemoryGithub_Submission/8_aPDL1scRNATCR"
so <-  paste0(path, "/data/8B_aPDL1_integratedseuratobj.rds") %>% readRDS() 
```

# set plot aesthetics
```{r}
# palette
clusterpal <- c(
  "Tcm" = "#65c7ca",
  "Tem_1" = "#f8a523",
  "Tem_2" = "#f6d13a",
  "Tex-Term_1" = "#bf3619",
  "Tex-Term_2" = "#7d1d19",
  "Tex-Int" = "#de311e",
  "Tex-KLR" = "#E27E73",
  "Tex-ISG" = "#a0a1d1",
  "Tex-Prog" = "#a94d9a",
  "Prolif" = "#e6af7c"
  )

clone.pal <- c("Divergent" = "#e6af7c", 
               "Tcm Bias" = "#65c7ca",
               "Tex-Term Bias" = "#bf3619",
               "Tex-Prog Bias" = "#a94d9a", 
               "Tem_1 Bias" = "#f8a523", 
               "Tem_2 Bias" = "#f6d13a"
                 )

group.pal <- c("PBS" = "#886231", "aPDL1" = "#299093")

plot.theme.umap <- theme_cowplot() + theme(plot.title = element_text(hjust = 0.5), 
        #panel.border = element_rect(colour = "black", fill=NA, linewidth = 0.5, color = "grey80"),
        axis.text = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5, face = "plain")) 

plot.theme <- theme_cowplot() + theme(plot.title = element_text(hjust = 0.5, face = "plain"))
```

# remove genes function
```{r}
#remove TCR genes from VF lists
remove_genes <- function(features, species, tcr=TRUE, ig=TRUE, cell_cycle=TRUE, mito=TRUE, histone=TRUE, ribosome=TRUE) {
    print(paste("Before filtering:", length(features), "features"))

    if (species == "human") {
        if (tcr) {
            features <- features[!grepl("^TR[ABGD][VDJC]", features)]
        }
        if (ig) {
            features <- features[!grepl("^IG[HLK][VDJC]", features)]
        }
        if (cell_cycle) {
            features <- features[!(features %in% unlist(cc.genes))]
        }
        if (mito) {
            features <- features[!grepl("^MT-", features)]
        }
    }
    if (species == "mouse") {
        if (tcr) {
            features <- features[!grepl("^Tr[abgd][vdjc]", features)]
        }
        if (tcr) {
            features <- features[!grepl("^Tcr[abgd][vdjc]", features)]
        }
        if (ig) {
            features <- features[!grepl("^Ig[hlk][vdjc]", features)]
        }
        if (cell_cycle) {
            features <- features[!(features %in% stringr::str_to_title(unlist(cc.genes)))]
        }
        if (mito) {
            features <- features[!grepl("^Mt-", features)]
        }
      
       if (histone) {
            features <- features[!grepl("^Hist", features)]
       }
      
       if (ribosome) {
            features <- features[!grepl("^Rps", features)]
       }
      
      if (ribosome) {
            features <- features[!grepl("^Rpl", features)]
        }
      
    }

    print(paste("After filtering:", length(features), "features"))
    
    return(features)
}
```

# find variable features, and remove TCR, mito, cell_clycle, histone, ribosome genes
```{r}

# variable features were run already, just doing it again to be safe 
DefaultAssay(so) <- "integrated"
so <- FindVariableFeatures(so)
vf.clean <- remove_genes(VariableFeatures(so), 
                         species = "mouse" , tcr=T, ig=F, cell_cycle=T, mito=T, histone=T, ribosome=T)
```

# find DEG of clusters
```{r}
DefaultAssay(so) <- "RNA"
all.markers <- FindAllMarkers(so, features = vf.clean, only.pos = T, logfc.threshold = 0.25, assay = "RNA", test.use = "wilcox", verbose = T)

# write.csv(all.markers, file = paste0(outs, "/3F - DEG per cluster.csv"))
# write_xlsx(all.markers, paste0(outs, "/3F - DEG per cluster.xlsx"))
```

# plot DEG
```{r}
# all deg per cluster ++++++++++++++++++++++++++++++++++++++++++++++++++++++++

hm.df <- DotPlot(so, features = unique(all.markers$gene))$data %>% 
  select(c(avg.exp, features.plot, id)) %>% 
  pivot_wider(names_from = id, values_from = avg.exp) %>% 
  as.data.frame() %>%  `rownames<-`(.[,1]) %>% select(-features.plot) %>%
  t() %>% scale()

## annotations

mark.df <- data.frame('gene' = colnames(hm.df), 'n' = 1:length(colnames(hm.df))) 
genes <- c("Tox", "Pdcd1", "Tim3", "Havcr2", "Lag3", "Cd101", "Sell", "Il7r", "Ccr7", "Tcf7", "Myb", "Slamf6", 
           "Cx3cr1", "Zeb2", "Klrg1", "Xcl1", "Gzmb","Isg15", "Id3", "Mki67", "Cxcr6", "Ifng", "Pclaf", "Cdkn3", "Cd7", "Cd160", "Ifit3", "Ifi1","Ifi27l2a", "Cxcr3", "Lgals3", "S1pr5")

gene.annot <- columnAnnotation(anno = anno_mark(at = mark.df$n[mark.df$gene %in% genes], 
                                             labels = mark.df$gene[mark.df$gene %in% genes], 
                                             which = "row", padding = 0.8))

ha.clust <- rowAnnotation(`Cluster` = rownames(hm.df), 
                        col = list(`Cluster` = clusterpal),
                        show_legend = F)

# plot HM
hm.alldeg <- hm.df %>% 
  Heatmap(name = "Z-score", 
          left_annotation = ha.clust, top_annotation = gene.annot,
          col = viridis(n = 9, option = "magma") ,
          # col = zrev(brewer.pal(9, "RdBu")), 
          width = unit(5, "in"), height = unit(2.2, "in"), 
          show_column_dend = F, show_column_names = F, row_dend_width = unit(0.2, "cm"),
          cluster_rows = T, cluster_columns = T, 
          border = T, column_names_rot = 45
        )

hm.alldeg

```

# Determine cluster proportion per group
```{r}
DimPlot(so, cols = clusterpal) +  plot.theme.umap 

# per group plots - UMAP ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

umap.df <- data.frame(so@meta.data,  so@reductions$umap@cell.embeddings)

# vector of groups
var.list <- c("PBS", "aPDL1") 

plots <- lapply(var.list, function(x) {
subset <- umap.df %>% subset(Group2 == x)
  umap <- ggplot(umap.df, aes(x = UMAP_1, y = UMAP_2)) +
    geom_point(color = "grey90", size = 0.3) +
    geom_point(data = subset, size = 0.3, aes(color = cell_cluster)) + 
    ggtitle(x) + 
    scale_color_manual(values = clusterpal)  + plot.theme.umap + NoLegend()
  
  bar <- subset %>% dplyr::count(cell_cluster, Group2) %>% mutate(n = n/sum(n)) %>%
    ggplot(aes(y = n, fill = cell_cluster, x = Group2)) + geom_col(show.legend = F) + 
    plot.theme + 
    labs(x = element_blank(), y = "Fraction of Cells") + 
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.02)), breaks=c(0, 0.5, 1)) + 
    scale_fill_manual(values = clusterpal) + coord_flip()
  
   plot_grid(umap, bar, rel_heights = c(1, 0.3), ncol = 1) %>% return()
   
})

do.call(plot_grid, c(plots, ncol = 2))


# per per mouse cluster dist ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

data.frame(so@meta.data,  so@reductions$umap@cell.embeddings) %>% 
  count(Group2, Rep, cell_cluster) %>% 
  filter(Group2 %in% c("PBS", "aPDL1")) %>% 
  mutate(Group2 = factor(Group2, levels = c("PBS", "aPDL1"))) %>%
  group_by(Group2, Rep) %>% 
  mutate(n = n/sum(n)) %>% 
  ggplot(aes(x = Group2, y = n)) +
    facet_wrap(~cell_cluster, scales = "free", ncol = 10) + 
    geom_boxplot(color = "black", outlier.shape = NA) + 
    geom_jitter(size = 2, width = 0.2, height = 0, aes(color = Group2)) + 
    plot.theme + 
    scale_color_manual(values = group.pal) + 
    theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(), strip.background = element_rect(fill = "transparent"), legend.position = "none") + 
    stat_compare_means(comparisons = list(c("aPDL1", "PBS")), tip.length = 0) +
    scale_y_continuous(expand = expansion(mult = c(0.1, 0.2))) + 
    labs(x = "Group", y = "Fraction of Cells")


```

# determine behaviors of clones in each group
```{r}
so@meta.data$clone_id <- paste(so$Group2, so$Rep, so$cdr3s_nt, sep = "__")

exp.clones <- so@meta.data %>% filter(Group2 %in% c("PBS", "aPDL1"), chains =="ab") %>% 
  dplyr::count(clone_id, name = "CloneSize") %>% 
  mutate(CloneFreq = CloneSize / sum(CloneSize)) %>% ungroup %>% 
  select(clone_id, CloneFreq, CloneSize) %>% filter(CloneSize > 2)

# phenotype proportion per clone
clone.df <- so@meta.data %>% filter(clone_id %in% exp.clones$clone_id) %>% 
  dplyr::count(cell_cluster, clone_id) %>%
  group_by(clone_id) %>% mutate(n = n/sum(n)) %>%
  pivot_wider(names_from = cell_cluster, values_from = n) %>%
  ungroup() %>% as.data.frame() %>%
  mutate(across(everything(), ~replace_na(.x, 0))) %>% 
  left_join(exp.clones) 
clone.df

clone.df$CloneSize %>% median

clone.df %>% select(-c(clone_id, CloneSize, CloneFreq, CloneSize)) %>% Heatmap(show_column_names = T, border = T, col= brewer.pal(9, "Purples"))

# cluster clones based on clone proportions
set.seed(22)
clusters <- kmeans(clone.df %>% select(-c(clone_id, CloneSize, CloneFreq)), centers = 8)
clone.df$Clone_Behavior <- clusters$cluster

# rename clusters
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "1"] <- "Tem_2 Bias"
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "2"] <- "Divergent"
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "3"] <- "Tex-Term Bias"
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "4"] <- "Tem_1 Bias"
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "5"] <- "Tem_1 Bias"
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "6"] <- "Tem_1 Bias"
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "7"] <- "Tcm Bias"
clone.df$Clone_Behavior[clone.df$Clone_Behavior == "8"] <- "Tex-Prog Bias"

# heatmaps
col.annot <- columnAnnotation(` `  = select(clone.df, -c(clone_id, Clone_Behavior, CloneFreq,  CloneSize)) %>% colnames(), 
              col = list(` ` = clusterpal), show_legend = F)


groups <- clone.df %>% separate(clone_id, into = c("group", "rep" ,"cdr3")) %>% select(group) %>% c()

size.annot <- rowAnnotation(Group = groups$group,
                            col = list(Group = group.pal), show_legend = T)

clone.hm <- clone.df %>% select(-c(clone_id, Clone_Behavior, CloneSize, CloneFreq)) %>% 
  Heatmap(name = "Fraction of Clone",
          show_column_names = T, border = T, col= brewer.pal(9, "Purples"), 
          row_split = clone.df$Clone_Behavior,
          show_row_dend = F, show_column_dend = T,
          row_title_rot = 0, 
          width = unit(10*0.6, "cm"), height = unit(8, "cm"), column_dend_height = unit(0.2, "cm"),
          top_annotation = col.annot, right_annotation = size.annot,
          column_title = paste0("n=", clone.df$clone_id %>% length(), " Clones >2 Cells") 
          )

# pdf(file = paste0(outs, "/3F_CloneHM_Cl13mem.pdf"))
clone.hm
# dev.off()


clone.df %>% separate(clone_id, into = c("group", "rep" , "cdr3")) %>% count(group, rep, Clone_Behavior) %>% 
  group_by(group, rep) %>% mutate(n = n/sum(n)) %>% 
  ggplot(aes(x = group, y = n)) + geom_boxplot() + geom_jitter(aes(color = group)) + facet_wrap(~Clone_Behavior) + stat_compare_means()
```


# PCA similarities between clusters between gorups
### to look at differences between clusters, we extract PC embeddings, then look at correlations of the top 30 PCs between groups
```{r}
# extract embedings and metadata
pca.df <- data.frame(so@meta.data %>% select(cell_cluster, Group2), Embeddings(so, reduction = "pca"))  %>% 
  filter(Group2 %in% c("PBS", "aPDL1")) %>%
  mutate(Pheno_Group = paste(Group2, cell_cluster, sep = "__"))

pca.df %>% head

# group by PhenoGroup and summarized PCs
summarized.pcs <- pca.df %>% group_by(Pheno_Group) %>% 
  dplyr::summarise(across(where(is.numeric), median, na.rm = TRUE))

# find correlations
cors <- summarized.pcs %>% column_to_rownames("Pheno_Group") %>% t() %>% Hmisc::rcorr() 

# select only correlations of interest, summarize adjusted p values
r.values <- cors$r[1:10,11:20] 
p.values <- cors$P[1:10,11:20]

sig_matrix <- matrix(" " , nrow = nrow(r.values), ncol = ncol(r.values))
colnames(sig_matrix) <- colnames(r.values)
rownames(sig_matrix) <- rownames(r.values)

# sumamrize bonferoni corrected pvalues
sig_matrix[p.values <=  0.05 / p.values %>% length()] <- "*" 
sig_matrix[p.values <=  0.01 / p.values %>% length()] <- "**" 
sig_matrix[p.values <=  0.001 / p.values %>% length()] <- "***"

# rename row and column names
rownames(r.values) <- gsub(".*__", "", rownames(r.values))
colnames(r.values) <- gsub(".*__", "", colnames(r.values))

# plot
col.annot <- columnAnnotation(` `  = colnames(r.values), col = list(` ` = clusterpal), show_legend = F)
row.annot <- rowAnnotation(` `  = colnames(r.values), col = list(` ` = clusterpal), show_legend = F)

hm.cor <- r.values %>% Heatmap(
                      col = viridis(9) , border = T, 
                      width = unit(7, "cm"), height = unit(7, "cm"), name = "Pearson R", 
                      cluster_rows = T, cluster_columns = T, row_dend_width = unit(0.2, "cm"), column_dend_height = unit(0.2, "cm"), row_dend_side = "right" ,
                      row_title = "aPDL1 Cluster" , column_title = "PBS Cluster" , 
                      row_title_side = "left", column_title_side = "bottom", row_names_side = "left",
                      left_annotation = row.annot, bottom_annotation = col.annot,
                      cell_fun = function(j, i, x, y, width, height, fill) {grid.text(sig_matrix[i, j], x, y, gp = gpar(fontsize = 8, col = "black"))},
                  )

hm.cor
```

# follow-up: DEG between cells in a given cluster between PBS and aPDL1 group
```{r}
test.list <- unique(so$cell_cluster)

# caclualte within cluster DEG
intra.cellular.deg <- lapply(test.list, function(x){

    ## subset only cells in cluster
    so.subset <- subset(so, cell_cluster == x)
    
    ## vector of up cells and down cells
    cells.up <- subset(so.subset, Group2 == "aPDL1") %>% colnames()
    cells.down <- subset(so.subset, Group2 == "PBS") %>% colnames()
    
    deg <- FindMarkers(object = so.subset, 
                       ident.1 = cells.up, ident.2 = cells.down, 
                       logfc.threshold = 0, assay = "RNA", features = vf.clean)
    
    deg <- deg %>% mutate(gene = rownames(deg)) %>%
      mutate(sig = dplyr::if_else(p_val_adj < 0.01 & avg_log2FC > 0.25, "aPDL1", "ns")) %>% 
      mutate(sig = dplyr::if_else(p_val_adj < 0.01 & avg_log2FC < -0.25,  "PBS", sig)) %>%
      filter(sig != "ns") %>% 
      mutate(cluster = x)
    
    return(deg)

})

# print number of DEG
do.call(rbind, intra.cellular.deg) %>% count(cluster, name = "aPDL1 vs PBS")

# calcualte DEG that define a cluster as a control
all.markers.updown <- FindAllMarkers(so, assay = "RNA", logfc.threshold = 0.25, only.pos = F, features = vf.clean)

# bind together and plot
merge(x = all.markers.updown %>% mutate(sig = dplyr::if_else(p_val_adj < 0.01, "sig", "ns")) %>% filter(sig != "ns") %>% count(cluster, name = "Cluster DEG") , 
     y = do.call(rbind, intra.cellular.deg) %>% count(cluster, name = "aPDL1 vs PBS") 
    ) %>% 
  pivot_longer(cols = -cluster) %>% 
  ggplot(aes(y = cluster, x = value, fill = name)) + geom_col(position = "dodge") +
  labs(x = '# of DEG', y = "Cluster") +
  plot.theme +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) + 
  scale_fill_manual(values = c("grey", "black"), name = element_blank()) 


```

