# Assess DEGs between clusters and look at cluster distribtuion per group

# load common libraries
```{r}
suppressPackageStartupMessages({
  library(tidyverse) 
  library(Seurat)
  library(RColorBrewer)
  library(viridis)
  library(cowplot)
  library(ggridges)
  library(data.table)
  library(ggpubr)
  library(ComplexHeatmap)
  library(readxl)
  library(writexl)
})
```

# load SO + set paths
```{r}
path <- "~/ChronicMemoryGithub_Submission/7_PolyclonalRechallenge/data/"
outs <- paste0(path, "/2_GEX_Analysis/outs")

so <- paste0(path, "7B_SeuratObject_Clustered.rds") %>% readRDS()

```

# set plot themes and color palettes
```{r}
plot.theme.umap <- theme_cowplot() + theme(plot.title = element_text(hjust = 0.5), 
        #panel.border = element_rect(colour = "black", fill=NA, linewidth = 0.5, color = "grey80"),
        axis.text = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) + 
  theme(plot.title = element_text(hjust = 0.5, face = "plain")) 

plot.theme <- theme_cowplot() + theme(plot.title = element_text(hjust = 0.5, face = "plain"))

clusterpal <- c(   
 
  "SLEC_1" = "#014f63",
  "SLEC_2" = "#62939a",
  "Eff_Mem" = "#65c7ca",
  "MPEC" = "#122452",
  
  "Exh_Eff" = "#ef8577",
  "Exh_EM" = "#e62230",
  "Exh_Term" = "#a11822",

  "Prolif_1" = "#f6d13a",    
  "Prolif_2" = "#f26e36",   
  "ISG" = "#e6af7c"
  )
```

# remove genes function
```{r}
#function to remove TCR genes from VF lists
remove_genes <- function(features, species, tcr=TRUE, ig=TRUE, cell_cycle=TRUE, mito=TRUE, histone=TRUE, ribosome=TRUE) {
    print(paste("Before filtering:", length(features), "features"))

    if (species == "human") {
        if (tcr) {
            features <- features[!grepl("^TR[ABGD][VDJC]", features)]
        }
        if (ig) {
            features <- features[!grepl("^IG[HLK][VDJC]", features)]
        }
        if (cell_cycle) {
            features <- features[!(features %in% unlist(cc.genes))]
        }
        if (mito) {
            features <- features[!grepl("^MT-", features)]
        }
    }
    if (species == "mouse") {
        if (tcr) {
            features <- features[!grepl("^Tr[abgd][vdjc]", features)]
        }
        if (tcr) {
            features <- features[!grepl("^Tcr[abgd][vdjc]", features)]
        }
        if (ig) {
            features <- features[!grepl("^Ig[hlk][vdjc]", features)]
        }
        if (cell_cycle) {
            features <- features[!(features %in% stringr::str_to_title(unlist(cc.genes)))]
        }
        if (mito) {
            features <- features[!grepl("^Mt-", features)]
        }
      
       if (histone) {
            features <- features[!grepl("^Hist", features)]
       }
      
       if (ribosome) {
            features <- features[!grepl("^Rps", features)]
       }
      
      if (ribosome) {
            features <- features[!grepl("^Rpl", features)]
        }
      
    }

    print(paste("After filtering:", length(features), "features"))
    
    return(features)
}
```

# find vf in SO and remove TCR and proliferation / histone / ribosome / mito related genes
```{r}
## this was the set of genes used to create PCs for clustering etc
DefaultAssay(so) <- "integrated"
vf.clean <- remove_genes(VariableFeatures(so), species = "mouse" , tcr=T, ig=F, cell_cycle=T, mito=T, histone=T, ribosome=T)
DefaultAssay(so) <- "RNA"
```

# ID prolif clusters
```{r}
DimPlot(so, label = T)

DimPlot(so, group.by = "Phase") & NoAxes() 
VlnPlot(so, features = c("S.Score", "G2M.Score"), pt.size = 0)
```

# find DEG per cluster
```{r}
# diferential genes per cluster
cluster.markers <- FindAllMarkers(so, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25, assay = "RNA", features = vf.clean)

cluster.markers$cell_cluster[cluster.markers$cluster == "0"] <- "SLEC_1"  # Called Teff in the manuscript
cluster.markers$cell_cluster[cluster.markers$cluster == "1"] <- "Exh_Eff"
cluster.markers$cell_cluster[cluster.markers$cluster == "2"] <- "Exh_EM" 
cluster.markers$cell_cluster[cluster.markers$cluster == "3"] <- "SLEC_2"  # Called Teff in the manuscript
cluster.markers$cell_cluster[cluster.markers$cluster == "4"] <- "Eff_Mem"
cluster.markers$cell_cluster[cluster.markers$cluster == "5"] <- "Prolif_1"
cluster.markers$cell_cluster[cluster.markers$cluster == "6"] <- "ISG"
cluster.markers$cell_cluster[cluster.markers$cluster == "7"] <- "Exh_Term"
cluster.markers$cell_cluster[cluster.markers$cluster == "8"] <- "Prolif_2"
cluster.markers$cell_cluster[cluster.markers$cluster == "9"] <- "MPEC"

# write_csv(cluster.markers, file = paste0(outs, "/2C_DegPerCluster.csv"))


```


# rename clusters
```{r}
DimPlot(so, label = T)

so@meta.data$cell_cluster[so@meta.data$seurat_clusters == "0"] <- "SLEC_1"  # Called Teff in the manuscript
so@meta.data$cell_cluster[so@meta.data$seurat_clusters == "1"] <- "Exh_Eff"
so@meta.data$cell_cluster[so@meta.data$seurat_clusters == "2"] <- "Exh_EM"
so@meta.data$cell_cluster[so@meta.data$seurat_clusters == "3"] <- "SLEC_2"  # Called Teff in the manuscript
so@meta.data$cell_cluster[so@meta.data$seurat_clusters == "4"] <- "Eff_Mem"
so@meta.data$cell_cluster[so@meta.data$seurat_clusters == "5"] <- "Prolif_1"
so@meta.data$cell_cluster[so@meta.data$seurat_clusters == "6"] <- "ISG"
so@meta.data$cell_cluster[so@meta.data$seurat_clusters == "7"] <- "Exh_Term"
so@meta.data$cell_cluster[so@meta.data$seurat_clusters == "8"] <- "Prolif_2"
so@meta.data$cell_cluster[so@meta.data$seurat_clusters == "9"] <- "MPEC"

so$cell_cluster <- factor(so$cell_cluster, levels = names(clusterpal))
Idents(so) <- so$cell_cluster
DimPlot(so,cols = clusterpal)

```

# dim reduction plots - nicer
```{r}
DimPlot(so, cols = clusterpal) + #+ plot.theme.umap + ggtitle(label = "CD8 T cells") 
  geom_text(data = data.frame(x = 1, y = 2), x = -6, y = -6, label = paste("n =", length(so$barcode), "cells"), size = 4) + 
  plot.theme.umap 

# dataframe for other UMAP based plots
umap.df <- data.frame(so@meta.data,  so@reductions$umap@cell.embeddings) %>% 
  mutate(Group = ifelse(Group == "AcuMem", "Arm:Arm", Group), 
         Group = ifelse(Group == "ChrMem", "Cl13:Arm", Group),
         Group = ifelse(Group == "EndoGP33", "Endo GP33+", Group))

# dim recution plots - split by Group =========================
head(umap.df)
# vector of groups
var.list <- c("Arm:Arm" ,"Cl13:Arm", "Endo GP33+") 

plots <- lapply(var.list, function(x) {
  subset <- umap.df %>% subset(Group == x)
  umap <- ggplot(umap.df, aes(x = UMAP_1, y = UMAP_2)) +
    ggrastr::geom_point_rast(color = "grey90", size = 0.1, raster.dpi = 600) +
    ggrastr::geom_point_rast(data = subset, size = 0.1, aes(color = cell_cluster), raster.dpi = 600) + 
    ggtitle(x) + 
    scale_color_manual(values = clusterpal)  + plot.theme.umap + NoLegend() 
  
   bar <- subset %>% dplyr::count(cell_cluster, Group) %>% mutate(n = n/sum(n)) %>%
    ggplot(aes(y = n, fill = cell_cluster, x = Group)) + geom_col(show.legend = F) + 
    plot.theme + 
    labs(x = element_blank(), y = "Fraction of Cells") + 
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.02)), breaks=c(0, 0.5, 1)) + 
    scale_fill_manual(values = clusterpal) + coord_flip()
  
   plot_grid(umap, bar, rel_heights = c(1, 0.3), ncol = 1) %>% return()

})

do.call(plot_grid, c(plots, ncol = 2))



# plot out frequencies for all samples
umap.df %>% count(Group, Tet, Rep, cell_cluster) %>% 
  mutate(Tet = ifelse(Tet == "GP33pos", "GP33+", "GP33-")) %>%
  mutate(Mouse = paste(Group, Tet, Rep, sep = "_")) %>% 
  mutate(Mouse = ifelse(Mouse == "Endo GP33+_GP33+_EndoGP33", "Endo GP33+", Mouse)) %>%
  group_by(Mouse) %>% mutate(n = n/sum(n)) %>% 
  ggplot(aes(x = Mouse, y = 100*n, fill = cell_cluster)) + 
  geom_col(show.legend = F) + 
  plot.theme + #rotate_x_text(90) +
  scale_fill_manual(values = clusterpal) + 
  labs(x = element_blank(), y = "% of Cells") +
  scale_y_continuous(expand = expansion(mult = c(0.0, 0.05))) + coord_flip()


```

# DEG plots
```{r}
# feature plots for well charactergorized genes
feat.plots <- FeaturePlot(so, 
                          features = c("Pdcd1", "Klrg1", "Cxcr3", "Tcf7"),
                          # cols =  c("grey", brewer.pal(4, "Reds")[c(2:4)]),
                          cols =  viridis(n = 9, option = "magma"),
                          order = T,
                          ncol = 5, pt.size = 0.01) & plot.theme.umap & NoLegend()

# legend
feat.legend <- cowplot::get_legend(FeaturePlot(so, features ="Cx3cr1", cols =  viridis(n = 9, option = "magma")))

plot_grid(feat.plots, feat.legend, rel_widths = c(13,1))

# ggsave(filename = "2C - Feature Plots.png", path = outs, height = 2.7, width = 10)
```

# top genes heatmap
```{r}
# get top 5 DEG per cluser
top5 <- cluster.markers %>% group_by(cluster) %>% slice_max(order_by = avg_log2FC, n = 5, with_ties = F)

hm.df <- DotPlot(so, features = unique(top5$gene))$data %>% 
  select(c(avg.exp, features.plot, id)) %>% 
  pivot_wider(names_from = id, values_from = avg.exp) %>% 
  as.data.frame() %>%  `rownames<-`(.[,1]) %>% select(-features.plot) %>%
  t() %>% scale()

ha.clust <- rowAnnotation(`Cluster` = rownames(hm.df), 
                        col = list(`Cluster` = clusterpal),
                        show_legend = F)

ha.clust

hm.top5 <- hm.df %>% 
  Heatmap(name = "Z-score", 
          right_annotation = ha.clust, 
          col = viridis(n = 9, option = "magma") ,
          # col = rev(brewer.pal(9, "RdYlBu")),
          width = unit(7, "in"), height = unit(2.2, "in"), 
          show_column_dend = F, show_row_names = T, row_dend_width = unit(0.15, "in"), 
          cluster_rows = T, cluster_columns = T, 
          border = T
        )

# pdf(file = paste0(outs, "/2C - HM Top5 DEG per cluster.pdf"), width = unit(12, "in"), height = unit(4, "in"))
hm.top5
# dev.off()
```

# add common exh. module score from Daniel Et al Nat Immuno (https://www.nature.com/articles/s41590-022-01337-5)
```{r}
# genes can be found in Supplemental tables of our manuscript
common.ex.genes <- read.csv("~/Resources/LCMV atlas tables/LCMV_common_ex_score.csv")$Common_Ex_Module
common.ex.genes <- common.ex.genes[common.ex.genes %in% vf.clean] %>% list()

DefaultAssay(so) <- "RNA"
so <- AddModuleScore(so, features = common.ex.genes, name = "CommonExMod", assay = "RNA")

VlnPlot(so, features = "CommonExMod1",  cols = clusterpal, pt.size = 0, sort = "decreasing") & plot.theme & rotate_x_text(45) & NoLegend()
FeaturePlot(so, features = "CommonExMod1")

common.ex.genes[[1]]

# heatmap of common Exh Genes
hm.df <- DotPlot(so, features = common.ex.genes[[1]])$data %>% 
  select(c(avg.exp, features.plot, id)) %>% 
  pivot_wider(names_from = id, values_from = avg.exp) %>% 
  as.data.frame() %>%  `rownames<-`(.[,1]) %>% select(-features.plot) %>%
  t() %>% scale()

ha.clust <- columnAnnotation(`Cluster` = rownames(hm.df), 
                        col = list(`Cluster` = clusterpal),
                        show_legend = F)

hm.df %>% t%>% 
  Heatmap(name = "Z-score", 
          top_annotation = ha.clust, 
          col = rev(brewer.pal(9, "RdYlBu")),
          width = unit(2, "in"), height = unit(4.7, "in"), 
          show_column_dend = F, show_row_names = T, row_dend_width = unit(0.15, "in"), 
          cluster_rows = T, cluster_columns = T, 
          border = T
        )
```

# also add module scores for Tex-Term, Teff, Tmem from  Daniel Et al Nat Immuno
```{r}
# read in genes from module scores (Daniel et al Table S1)
## we only use the effector module in our analysis - see our table S7
genes.tables <- read_xls("~/Resources/LCMV atlas tables/Supplemental Table 1.xls") 

# take top 50 genes fgrom each 
term.genes <- genes.tables %>% filter(cluster == "Texterm") %>% slice_max(n = 50, order_by = avg_log2FC) %>% select(gene) %>% c() 
mem.genes <- genes.tables %>% filter(cluster == "Tmem") %>% slice_max(n = 50, order_by = avg_log2FC) %>% select(gene) %>% c() 
eff.genes <- genes.tables %>% filter(cluster == "Teff") %>% slice_max(n = 50, order_by = avg_log2FC) %>% select(gene) %>% c() 

# filter gby 
module.lists <- list(
  "Term_Module" = term.genes$gene[term.genes$gene %in% vf.clean],
  "Mem_Module" = mem.genes$gene[mem.genes$gene %in% vf.clean],
  "Teff_Module" = eff.genes$gene[eff.genes$gene %in% vf.clean]
)

so <- AddModuleScore(so, features = module.lists, name = names(module.lists), replace = T, assay = "RNA")

# plot out Moldule scores
VlnPlot(so, features = c("Term_Module1", "Mem_Module2", "Teff_Module3"),  cols = clusterpal, pt.size = 0, sort = "decreasing", ncol = 3) & 
  plot.theme & rotate_x_text(90) & NoLegend()
# ggsave(filename = "2B - ModuleScoreVlns.pdf", path = outs, height = 9, width = 3)

# print out mopdules
print("Tmem")
module.lists$Mem_Module
print("Teff")
module.lists$Teff_Module
print("Tex")
module.lists$Term_Module

```

# after adding module scores, save RDS
```{r}
# saveRDS(so, file = paste0(path, "/seurat_objects/7C_SeuratObject_ClusteredAndIDd.rds"))
```


